---
title: "PRJNA766024 Analysis"
output:
  html_document:
    df_print: paged
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
source(here::here("src", "helpers", "helpers.R"))
```

## Load Libraries

```{r}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(coriell))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(MultiAssayExperiment))
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(qsmooth))
suppressPackageStartupMessages(library(quantro))
suppressPackageStartupMessages(library(doParallel))
suppressPackageStartupMessages(library(PCAtools))
suppressPackageStartupMessages(library(patchwork))
suppressPackageStartupMessages(library(paletteer))
suppressPackageStartupMessages(library(rms))
suppressPackageStartupMessages(library(Hmisc))
suppressPackageStartupMessages(library(ggbeeswarm))
registerDoParallel(cores = 12)
```

## Read in MultiAssayExperiment Object

```{r}
mae <- readRDS(here("data", "PRJNA766024", "MultiAssayExperiment.rds"))
cds <- readRDS(here("doc", "gencode.v26.annotation.cds.rds"))
genes <- assay(mae, "gene counts")
repeats <- assay(mae, "RE counts")
repeats <- subset(repeats, grepl(rownames(repeats), pattern = "^LINE\\.|^SINE\\.|^LTR\\.|.*\\.SVA\\..*|^DNA\\."))
genes <- subset(genes, rownames(genes) %in% cds$gene_name)

# check for consistent colnames
all(colnames(genes) == colnames(repeats))

# bind both matrices into single count matrix
counts <- rbind(genes, repeats)
counts <- sumTechReps(counts, ID = colData(mae)$BioSample)
colData(mae)$bases <- as.integer(colData(mae)$bases)
metadata <- as.data.table(colData(mae))
metadata <- unique(metadata[, .(BioSample, group)])
metadata[, c("cell_line", "treatment") := tstrsplit(group, ".", fixed = TRUE)]
setDF(metadata, rownames = metadata$BioSample)
```

## Check for global differences between groups with `quantro`

```{r}
qtest <- quantro(counts, groupFactor = metadata$group, B = 1e3)
perm_pval <- quantroPvalPerm(qtest)
```

## DE pipeline

```{r}
# Define experimental design
design <- model.matrix(~0 + group, data = metadata)
colnames(design) <- gsub(pattern = "^group", replacement = "", x = colnames(design))

# Define contrasts to test
contrast.matrix <- makeContrasts(
  A375.THZ531_500nM_6H_vs_DMSO = A375.THZ531_500nM_6H - A375.DMSO_6H,
  Colo829.THZ531_500nM_6H_vs_DMSO = Colo829.THZ531_500nM_6H - Colo829.DMSO_6H,
  levels = design
)

# Create DGEList object from counts and filter for expression
y <- DGEList(counts = counts, samples = metadata)
keep <- filterByExpr(y, design = design)
y <- y[keep,, keep.lib.sizes = FALSE]

# If global differences exist use QSmooth normalization, if not use TMM
if (perm_pval < 0.01) {
  norm_method <- "qs"
  qs <- qsmooth(y$counts, group_factor = y$samples$group)
  fit <- lmFit(log2(qsmoothData(qs) + 2), design)
  fit <- contrasts.fit(fit, contrast.matrix)
} else {
  norm_method <- "tmm"
  y <- calcNormFactors(y, method = "TMM")
  y <- estimateDisp(y, design, robust = TRUE)
  fit <- glmQLFit(y, design, robust = TRUE)
  plotQLDisp(fit)
}
```

## Test for differential expression above a fold-change

Use either `treat` or `glmTreat` depending on the fit type to test for 
differential expression above a fold-change.

```{r}
# Define fold-change that features must exceed to be considered for significance
FC <- 1.5

contrast_names <- colnames(contrast.matrix)
if (norm_method == "qs") {
  fit2 <- treat(fit, lfc = log2(FC), trend = TRUE, robust = TRUE)
  plotSA(fit2)
  res_dts <- get_treat(contrast_names, fit2)
} else {
  res_dts <- get_glmTreat(contrast_names, fit, contrast.matrix, FC)
  }

# extract all results and bind to single df
df <- rbindlist(res_dts, idcol = "contrast")

# Calculate logcounts of filtered, normalized data
if (norm_method == "qs") {
  logcounts <- log2(qsmoothData(qs) + 2)
} else {
  logcounts <- cpm(y, log = TRUE, prior.count = 2)
}

# Subset for genes and REs
logcounts.genes <- logcounts[rownames(logcounts) %in% rownames(genes), ]
logcounts.repeats <- logcounts[rownames(logcounts) %in% rownames(repeats), ]
```

## PCA

```{r}
pca.genes <- pca(logcounts.genes, metadata, scale = TRUE, center = TRUE)
pca.repeats <- pca(logcounts.repeats, metadata, scale = TRUE, center = TRUE)

b1 <- biplot(
  pca.genes, 
  colby = "treatment", 
  shape = "cell_line",
  legendPosition = "bottom", 
  lab = NULL, 
  pointSize = 5,
  hline = 0, 
  hlineType = 2,
  vline = 0, 
  vlineType = 2,
  title = "Protein Coding Genes"
  ) + 
  scale_color_manual(values = c("DMSO_6H" = "steelblue", "THZ531_500nM_6H" = "firebrick")) +
  theme_coriell()

b2 <- biplot(
  pca.repeats, 
  colby = "treatment", 
  shape = "cell_line",
  legendPosition = "bottom", 
  lab = NULL, 
  pointSize = 5,
  hline = 0, 
  hlineType = 2,
  vline = 0, 
  vlineType = 2,
  title = "Repetitive Elements"
  ) + 
  scale_color_manual(values = c("DMSO_6H" = "steelblue", "THZ531_500nM_6H" = "firebrick")) +
  theme_coriell()

(b1  | b2) + plot_layout(guides = "collect") & theme_coriell()
ggsave(here("results", "PRJNA766024", "figures", "biplots.png"), width = 12, height = 6)
```

## Heatmap

```{r}
ann_cols <- list(
  cell_line = c("A375" = "#0073C2", "Colo829" = "#EFC000"),
  treatment = c("DMSO_6H" = "steelblue", "THZ531_500nM_6H" = "firebrick")
  )

quickmap(
  logcounts.genes, 
  annotation_col = metadata[, c("treatment", "cell_line")],
  annotation_colors = ann_cols,
  show_colnames = FALSE,
  main = "Protein Coding Genes",
  filename = here("results", "PRJNA766024", "figures", "heatmap-genes.png")
  )

quickmap(
  logcounts.repeats, 
  annotation_col = metadata[, c("treatment", "cell_line")],
  annotation_colors = ann_cols, 
  show_colnames = FALSE,
  main = "Repetitive Elements",
  filename = here("results", "PRJNA766024", "figures", "heatmap-repeats.png")
  )
```

## Is differential expression related to gene length?

> We find that melanoma cells harbor constitutively high CDK12 activity, and that its inhibition decreases the expression 
of long genes containing multiple exons, including many genes involved in DNA repair. Conversely, our results show that
CDK12 inhibition promotes the expression of short genes with few exons, including many growth-promoting genes regulated 
by the AP-1 and NF-κB transcription factors

```{r}
# Import the gene reference
gtf <- rtracklayer::import("/mnt/data/gdata/human/REdiscoverTE_hg38/gencode.v26.annotation.gtf.gz")

# Filter for the genes in the count matrix
coding <- gtf[gtf$type == "gene" & gtf$gene_type == "protein_coding" & gtf$gene_name %in% rownames(y$counts), ]

# Get the size of each feature
widths <- data.table(gene_name = coding$gene_name, width = width(coding))

# Since some features (44) have two lengths, take the maximum
widths <- widths[, .(width = max(width)), by = gene_name]

# Join the gene lengths onto the DE results
df <- widths[df, on = c("gene_name" = "feature_id"), nomatch = NA]

# Label the features as either up or down-regulated
df[, direction := fcase(FDR < 0.1 & logFC > 0, "up",
                        FDR < 0.1 & logFC < 0, "down",
                        default = "unperturbed")][, 
     `:=`(direction = factor(direction, levels = c("unperturbed", "down", "up")),
          logWidth = log10(width))]
```

```{r}
v1 <- plot_volcano(
  subset(df, contrast == "A375.THZ531_500nM_6H_vs_DMSO" & gene_name %in% rownames(genes)), 
  lfc = log2(1.5)) + labs(title = "A375 THZ531 500nM 6H vs DMSO")
m1 <- plot_md(
  subset(df, contrast == "A375.THZ531_500nM_6H_vs_DMSO" & gene_name %in% rownames(genes)), lfc = log2(1.5), 
  x = "AveExpr") + labs(title = "A375 THZ531 500nM 6H vs DMSO")
v2 <- plot_volcano(
  subset(df, contrast == "Colo829.THZ531_500nM_6H_vs_DMSO" & gene_name %in% rownames(genes)), 
  lfc = log2(1.5)) + labs(title = "Colo829 THZ531 500nM 6H vs DMSO")
m2 <- plot_md(
  subset(df, contrast == "Colo829.THZ531_500nM_6H_vs_DMSO" & gene_name %in% rownames(genes)), lfc = log2(1.5), 
  x = "AveExpr") + labs(title = "Colo829 THZ531 500nM 6H vs DMSO")

(v1 | v2) / (m1 | m2) +
  plot_annotation(title = "Protein Coding Genes") & theme_coriell()
ggsave(here("results", "PRJNA766024", "figures", "diffExpr-gene.png"), width = 18, height = 12)

# repeats
v1 <- plot_volcano(
  subset(df, contrast == "A375.THZ531_500nM_6H_vs_DMSO" & !gene_name %in% rownames(genes)), 
  lfc = log2(1.5)) + labs(title = "A375 THZ531 500nM 6H vs DMSO")
m1 <- plot_md(
  subset(df, contrast == "A375.THZ531_500nM_6H_vs_DMSO" & !gene_name %in% rownames(genes)), lfc = log2(1.5), 
  x = "AveExpr") + labs(title = "A375 THZ531 500nM 6H vs DMSO")
v2 <- plot_volcano(
  subset(df, contrast == "Colo829.THZ531_500nM_6H_vs_DMSO" & !gene_name %in% rownames(genes)), 
  lfc = log2(1.5)) + labs(title = "Colo829 THZ531 500nM 6H vs DMSO")
m2 <- plot_md(
  subset(df, contrast == "Colo829.THZ531_500nM_6H_vs_DMSO" & !gene_name %in% rownames(genes)), lfc = log2(1.5), 
  x = "AveExpr") + labs(title = "Colo829 THZ531 500nM 6H vs DMSO")

(v1 | v2) / (m1 | m2) +
  plot_annotation(title = "Repetitive Elements") & theme_coriell()
ggsave(here("results", "PRJNA766024", "figures", "diffExpr-repeats.png"), width = 18, height = 12)
```

```{r}
# Subset for each separately
a375 <- df[contrast == "A375.THZ531_500nM_6H_vs_DMSO" & gene_name %in% rownames(genes)]
Colo829 <- df[contrast == "Colo829.THZ531_500nM_6H_vs_DMSO" & gene_name %in% rownames(genes)]
```

```{r}
# A375
p1 <- ggplot(a375, aes(x = width, y = logFC)) +
  geom_point(data = subset(a375, direction == "unperturbed"), color = "grey80", size = 0.5, alpha = 0.5) +
  geom_point(data = subset(a375, direction == "up"), color = "red2", size = 0.5, alpha = 0.8) +
  geom_point(data = subset(a375, direction == "down"), color = "blue2", size = 0.5, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_smooth(data = subset(a375, direction == "unperturbed"), method = "gam", se = FALSE, color = "grey65") +
  geom_smooth(data = subset(a375, direction == "up"), method = "gam", se = FALSE, color = "red3") +
  geom_smooth(data = subset(a375, direction == "down"), method = "gam", se = FALSE, color = "blue3") +
  geom_smooth(method = "gam", se = FALSE, color = "black", linetype = 2) +
  geom_rug(sides = "bl", alpha = 0.05) +
  scale_x_log10() +
  labs(title = "A375 THZ531 vs DMSO 6hr",
       x = "Gene Length (bp)",
       y = "logFC") +
  theme_coriell()

p2 <- ggplot(Colo829, aes(x = width, y = logFC)) +
  geom_point(data = subset(Colo829, direction == "unperturbed"), color = "grey80", size = 0.5, alpha = 0.5) +
  geom_point(data = subset(Colo829, direction == "up"), color = "red2", size = 0.5, alpha = 0.8) +
  geom_point(data = subset(Colo829, direction == "down"), color = "blue2", size = 0.5, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_smooth(data = subset(Colo829, direction == "unperturbed"), method = "gam", se = FALSE, color = "grey65") +
  geom_smooth(data = subset(Colo829, direction == "up"), method = "gam", se = FALSE, color = "red3") +
  geom_smooth(data = subset(Colo829, direction == "down"), method = "gam", se = FALSE, color = "blue3") +
  geom_smooth(method = "gam", se = FALSE, color = "black", linetype = 2) +
  geom_rug(sides = "bl", alpha = 0.05) +
  scale_x_log10() +
  labs(title = "Colo829 THZ531 vs DMSO 6hr",
       x = "Gene Length (bp)",
       y = "logFC") +
  theme_coriell()

(p1 | p2) +
  plot_annotation(title = "Gene Length (introns + exons)") & theme_coriell()
ggsave(here("results", "PRJNA766024", "figures", "width_vs_logFC.png"), width = 18, height = 8)
```

```{r}
p1 <- ggplot(a375, aes(x = width, y = logFC)) +
  geom_point(data = subset(a375, direction == "unperturbed"), color = "grey80", size = 0.5, alpha = 0.5) +
  geom_point(data = subset(a375, direction == "up"), aes(color = AveExpr), size = 2, alpha = 0.8) +
  geom_point(data = subset(a375, direction == "down"), aes(color = AveExpr), size = 2, alpha = 0.8) +
  geom_smooth(data = subset(a375, direction == "unperturbed"), method = "gam", se = FALSE, color = "grey65") +
  geom_smooth(data = subset(a375, direction == "up"), method = "gam", se = FALSE, color = "red3") +
  geom_smooth(data = subset(a375, direction == "down"), method = "gam", se = FALSE, color = "blue3") +
  geom_smooth(method = "gam", se = FALSE, color = "black", linetype = 2) +
  geom_rug(sides = "bl", alpha = 0.05) +
  scale_color_viridis_c(direction = -1) +
  scale_x_log10() +
  labs(title = "A375 THZ531 vs DMSO 6hr",
       x = "Gene Length (bp)",
       y = "logFC") +
  theme_coriell()

p2 <- ggplot(Colo829, aes(x = width, y = logFC)) +
  geom_point(data = subset(Colo829, direction == "unperturbed"), color = "grey80", size = 0.5, alpha = 0.5) +
  geom_point(data = subset(Colo829, direction == "up"), aes(color = AveExpr), size = 2, alpha = 0.8) +
  geom_point(data = subset(Colo829, direction == "down"), aes(color = AveExpr), size = 2, alpha = 0.8) +
  geom_smooth(data = subset(Colo829, direction == "unperturbed"), method = "gam", se = FALSE, color = "grey65") +
  geom_smooth(data = subset(Colo829, direction == "up"), method = "gam", se = FALSE, color = "red3") +
  geom_smooth(data = subset(Colo829, direction == "down"), method = "gam", se = FALSE, color = "blue3") +
  geom_smooth(method = "gam", se = FALSE, color = "black", linetype = 2) +
  geom_rug(sides = "bl", alpha = 0.05) +
  scale_color_viridis_c(direction = -1) +
  scale_x_log10() +
  labs(title = "Colo829 THZ531 vs DMSO 6hr",
       x = "Gene Length (bp)",
       y = "logFC") +
  theme_coriell()

(p1 | p2) +
  plot_annotation(title = "Gene Length vs Differential Expression") & theme_coriell()
ggsave(here("results", "PRJNA766024", "figures", "width_vs_logFC2.png"), width = 18, height = 8)
```

How strong is this association? -- it's not.

```{r}
d <- datadist(a375)
options(datadist = "d")
f <- ols(logFC ~ rcs(logWidth, 4) * direction, data = a375)
f
```

```{r}
d <- datadist(Colo829)
options(datadist = "d")
f <- ols(logFC ~ rcs(logWidth, 4) * direction, data = Colo829)
f
```

## By exon lengths

The quant files contain the effective lengths of each transcript. These are the length estimates adjusted for the 
relative probability of a read being align.

```{r}
# Get the tx2gene file that maps transcripts / REs to gene names
tx2gene <- readRDS("/mnt/data/gdata/human/REdiscoverTE_hg38/tx2gene_REdiscoverTE.rds")
setkey(tx2gene, "md5")

# Get the effective lengths for each transcript from the quant files
quant_files <- list.files(
  path = "/mnt/data/data_gc/projects/geo-rna-seq/data/PRJNA766024/quants",
  pattern = "quant.sf",
  recursive = TRUE,
  full.names = TRUE
  )
names(quant_files) <- regmatches(quant_files, regexpr("SRR[0-9]+", quant_files))

# Read all quant_files into a single dt
quant <- rbindlist(lapply(quant_files, fread, select = c("Name", "Length", "EffectiveLength")), idcol = "Run")

# Get the mean values for the lengths and EffectiveLengths of each transcript
by_feature <- quant[, .(Length = mean(Length), EffectiveLength = mean(EffectiveLength)), by = Name]
setkey(by_feature, "Name")

# Join the gene name information onto the summarized lengths
by_feature <- tx2gene[by_feature]
```

Sum the lengths and effective lengths for each gene

```{r}
a375_features <- by_feature[feature_id %chin% a375[, unique(gene_name)]][, 
                           .(sumLength = sum(Length), sumEffectiveLength = sum(EffectiveLength)), 
                           by = feature_id]
colo_features <- by_feature[feature_id %chin% Colo829[, unique(gene_name)]][, 
                           .(sumLength = sum(Length), sumEffectiveLength = sum(EffectiveLength)), 
                           by = feature_id]


# Join back onto the data 
a375 <- a375[a375_features, on = c("gene_name" = "feature_id")]
Colo829 <- Colo829[colo_features, on = c("gene_name" = "feature_id")]
```

re-plot using the sum of the transcripts and effective lengths

```{r}
p1 <- ggplot(a375, aes(x = sumLength, y = logFC)) +
  geom_point(data = subset(a375, direction == "unperturbed"), color = "grey80", size = 0.5, alpha = 0.5) +
  geom_point(data = subset(a375, direction == "up"), color = "red2", size = 0.5, alpha = 0.8) +
  geom_point(data = subset(a375, direction == "down"), color = "blue2", size = 0.5, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_smooth(data = subset(a375, direction == "unperturbed"), method = "gam", se = FALSE, color = "grey65") +
  geom_smooth(data = subset(a375, direction == "up"), method = "gam", se = FALSE, color = "red3") +
  geom_smooth(data = subset(a375, direction == "down"), method = "gam", se = FALSE, color = "blue3") +
  geom_smooth(method = "gam", se = FALSE, color = "black", linetype = 2) +
  geom_rug(sides = "bl", alpha = 0.05) +
  scale_x_log10() +
  labs(title = "A375 THZ531 vs DMSO 6hr",
       x = "Gene Length (bp)",
       y = "logFC") +
  theme_coriell()

p2 <- ggplot(Colo829, aes(x = sumLength, y = logFC)) +
  geom_point(data = subset(Colo829, direction == "unperturbed"), color = "grey80", size = 0.5, alpha = 0.5) +
  geom_point(data = subset(Colo829, direction == "up"), color = "red2", size = 0.5, alpha = 0.8) +
  geom_point(data = subset(Colo829, direction == "down"), color = "blue2", size = 0.5, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_smooth(data = subset(Colo829, direction == "unperturbed"), method = "gam", se = FALSE, color = "grey65") +
  geom_smooth(data = subset(Colo829, direction == "up"), method = "gam", se = FALSE, color = "red3") +
  geom_smooth(data = subset(Colo829, direction == "down"), method = "gam", se = FALSE, color = "blue3") +
  geom_smooth(method = "gam", se = FALSE, color = "black", linetype = 2) +
  geom_rug(sides = "bl", alpha = 0.05) +
  scale_x_log10() +
  labs(title = "Colo829 THZ531 vs DMSO 6hr",
       x = "Gene Length (bp)",
       y = "logFC") +
  theme_coriell()

(p1 | p2) +
  plot_annotation(title = "Sum Transcript Lengths per Gene") & theme_coriell()
ggsave(here("results", "PRJNA766024", "figures", "width_vs_logFC3.png"), width = 18, height = 8)
```

by Effective lengths

```{r}
p1 <- ggplot(a375, aes(x = sumEffectiveLength, y = logFC)) +
  geom_point(data = subset(a375, direction == "unperturbed"), color = "grey80", size = 0.5, alpha = 0.5) +
  geom_point(data = subset(a375, direction == "up"), color = "red2", size = 0.5, alpha = 0.8) +
  geom_point(data = subset(a375, direction == "down"), color = "blue2", size = 0.5, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_smooth(data = subset(a375, direction == "unperturbed"), method = "gam", se = FALSE, color = "grey65") +
  geom_smooth(data = subset(a375, direction == "up"), method = "gam", se = FALSE, color = "red3") +
  geom_smooth(data = subset(a375, direction == "down"), method = "gam", se = FALSE, color = "blue3") +
  geom_smooth(method = "gam", se = FALSE, color = "black", linetype = 2) +
  geom_rug(sides = "bl", alpha = 0.05) +
  scale_x_log10() +
  labs(title = "A375 THZ531 vs DMSO 6hr",
       x = "Gene Length (bp)",
       y = "logFC") +
  theme_coriell()

p2 <- ggplot(Colo829, aes(x = sumEffectiveLength, y = logFC)) +
  geom_point(data = subset(Colo829, direction == "unperturbed"), color = "grey80", size = 0.5, alpha = 0.5) +
  geom_point(data = subset(Colo829, direction == "up"), color = "red2", size = 0.5, alpha = 0.8) +
  geom_point(data = subset(Colo829, direction == "down"), color = "blue2", size = 0.5, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_smooth(data = subset(Colo829, direction == "unperturbed"), method = "gam", se = FALSE, color = "grey65") +
  geom_smooth(data = subset(Colo829, direction == "up"), method = "gam", se = FALSE, color = "red3") +
  geom_smooth(data = subset(Colo829, direction == "down"), method = "gam", se = FALSE, color = "blue3") +
  geom_smooth(method = "gam", se = FALSE, color = "black", linetype = 2) +
  geom_rug(sides = "bl", alpha = 0.05) +
  scale_x_log10() +
  labs(title = "Colo829 THZ531 vs DMSO 6hr",
       x = "Gene Length (bp)",
       y = "logFC") +
  theme_coriell()

(p1 | p2) +
  plot_annotation(title = "Sum Effective Lengths of Transcripts per Gene") & theme_coriell()
ggsave(here("results", "PRJNA766024", "figures", "width_vs_logFC4.png"), width = 18, height = 8)
```

### MA plot with the lengths

```{r}
v1 <- ggplot(a375, aes(x = logFC, y = -log10(FDR), color = log10(sumEffectiveLength))) +
  geom_point(alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = 2) +
  scale_color_viridis_c(direction = -1) +
  labs(title = "A375 THZ531 vs DMSO 6hr",
       x = "logFC",
       y = "-log10(FDR)",
       color = "log10(length)") +
  theme_coriell()

v2 <- ggplot(Colo829, aes(x = logFC, y = -log10(FDR), color = log10(sumEffectiveLength))) +
  geom_point(alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = 2) +
  scale_color_viridis_c(direction = -1) +
  labs(title = "Colo829 THZ531 vs DMSO 6hr",
       x = "logFC",
       y = "-log10(FDR)",
       color = "log10(length)") +
  theme_coriell()

m1 <- ggplot(a375, aes(x = AveExpr, y = logFC, color = log10(sumEffectiveLength))) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = 2) +
  scale_color_viridis_c(direction = -1) +
  scale_x_log10() +
  labs(title = "A375 THZ531 vs DMSO 6hr",
       x = "AveExpr",
       y = "logFC",
       color = "log10(length)") +
  theme_coriell()

m2 <- ggplot(Colo829, aes(x = AveExpr, y = logFC, color = log10(sumEffectiveLength))) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = 2) +
  scale_color_viridis_c(direction = -1) +
  scale_x_log10() +
  labs(title = "Colo829 THZ531 vs DMSO 6hr",
       x = "AveExpr",
       y = "logFC",
       color = "log10(length)") +
  theme_coriell()

(v1 | v2) / (m1 | m2) +
  plot_layout(guides = "collect") & 
  theme_coriell()
ggsave(here("results", "PRJNA766024", "figures", "diffExpr-gene-byLength.png"), width = 18, height = 12)
```

## What are the effective lengths of REs?

```{r}
re_lengths <- by_feature[feature_id %like% "__"]
re_lengths[, c("class", "family", "subfamily_loc") := tstrsplit(feature_id, ".", fixed = TRUE)][, feature_id := NULL]
re_lengths[, c("subfamily", "location") := tstrsplit(subfamily_loc, "__", fixed = TRUE)][, subfamily_loc := NULL]
re_lengths[, repElem := paste(class, family, subfamily, sep = ".")]
```

```{r}
re_summary <- re_lengths[repElem %like% "^LINE\\.|^SINE\\.|^LTR\\.|.*\\.SVA\\..*|^DNA\\.", 
           .(meanLength = mean(Length), meanEffectiveLength = mean(EffectiveLength),
             medianLength = median(Length), medianEffectiveLength = median(EffectiveLength),
             maxLength = max(Length), maxEffectiveLength = max(EffectiveLength),
             minLength = min(Length), minEffectiveLength = min(EffectiveLength)),
           by = .(class, family, subfamily)][order(meanLength, decreasing = TRUE)]
re_summary[, repElem := paste(class, family, subfamily, sep = ".")]
```

```{r}
p1 <- ggplot(re_summary, aes(x = reorder(class, medianEffectiveLength), y = medianEffectiveLength)) +
  geom_quasirandom(alpha = 0.5) +
  scale_y_log10() +
  labs(title = "Median Effective Length",
       x = "Class",
       y = "Median Effective Length"
       ) +
  theme_coriell()

p2 <- ggplot(re_summary, aes(x = reorder(class, meanEffectiveLength), y = meanEffectiveLength)) +
  geom_quasirandom(alpha = 0.5) +
  scale_y_log10() +
  labs(title = "Mean Effective Length",
       x = "Class",
       y = "Mean Effective Length"
       ) +
  theme_coriell()

p3 <- ggplot(re_summary, aes(x = reorder(class, medianLength), y = medianLength)) +
  geom_quasirandom(alpha = 0.5) +
  scale_y_log10() +
  labs(title = "Median Length",
       x = "Class",
       y = "Median Length"
       ) +
  theme_coriell()

p4 <- ggplot(re_summary, aes(x = reorder(class, meanLength), y = meanLength)) +
  geom_quasirandom(alpha = 0.5) +
  scale_y_log10() +
  labs(title = "Mean Length",
       x = "Class",
       y = "Mean Length"
       ) +
  theme_coriell()

(p1 | p2) / (p3 | p4) +
  plot_annotation(title = "Distributions of RE Lengths")
ggsave(here("results", "PRJNA766024", "figures", "repeat-length-distributions.png"), width = 12, height = 8)
```

Is effective length related to expression?

```{r}
df_re <- df[re_summary, on = c("gene_name" = "repElem"), nomatch = NULL]
a375_re <- df_re[contrast == "A375.THZ531_500nM_6H_vs_DMSO"]
Colo829_re <- df_re[contrast == "Colo829.THZ531_500nM_6H_vs_DMSO"]
```

```{r}
p1 <- ggplot(a375_re, aes(x = medianEffectiveLength, y = logFC)) +
  geom_point(data = subset(a375_re, direction == "unperturbed"), color = "grey80", size = 0.5, alpha = 0.5) +
  geom_point(data = subset(a375_re, direction == "up"), color = "red2", size = 0.5, alpha = 0.8) +
  geom_point(data = subset(a375_re, direction == "down"), color = "blue2", size = 0.5, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_smooth(data = subset(a375_re, direction == "unperturbed"), method = "gam", se = FALSE, color = "grey65") +
  geom_smooth(data = subset(a375_re, direction == "up"), method = "gam", se = FALSE, color = "red3") +
  geom_smooth(data = subset(a375_re, direction == "down"), method = "gam", se = FALSE, color = "blue3") +
  geom_smooth(method = "gam", se = FALSE, color = "black", linetype = 2) +
  geom_rug(sides = "bl", alpha = 0.05) +
  scale_x_log10() +
  labs(title = "A375 THZ531 vs DMSO 6hr",
       x = "median Effective Length",
       y = "logFC") +
  theme_coriell()

p2 <- ggplot(Colo829_re, aes(x = medianEffectiveLength, y = logFC)) +
  geom_point(data = subset(Colo829_re, direction == "unperturbed"), color = "grey80", size = 0.5, alpha = 0.5) +
  geom_point(data = subset(Colo829_re, direction == "up"), color = "red2", size = 0.5, alpha = 0.8) +
  geom_point(data = subset(Colo829_re, direction == "down"), color = "blue2", size = 0.5, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_smooth(data = subset(Colo829_re, direction == "unperturbed"), method = "gam", se = FALSE, color = "grey65") +
  geom_smooth(data = subset(Colo829_re, direction == "up"), method = "gam", se = FALSE, color = "red3") +
  geom_smooth(data = subset(Colo829_re, direction == "down"), method = "gam", se = FALSE, color = "blue3") +
  geom_smooth(method = "gam", se = FALSE, color = "black", linetype = 2) +
  geom_rug(sides = "bl", alpha = 0.05) +
  scale_x_log10() +
  labs(title = "Colo829 THZ531 vs DMSO 6hr",
       x = "median Effective Length",
       y = "logFC") +
  theme_coriell()

(p1 | p2) +
  plot_annotation(title = "Median Effective Length per RE") & theme_coriell()
ggsave(here("results", "PRJNA766024", "figures", "width_vs_logFC-re.png"), width = 18, height = 8)
```

```{r}
p1 <- ggplot(a375_re, aes(x = medianEffectiveLength, y = logFC)) +
  geom_point(data = subset(a375_re, direction == "unperturbed"), color = "grey80", size = 0.5, alpha = 0.5) +
  geom_point(data = subset(a375_re, direction == "up"), aes(color = AveExpr), size = 2, alpha = 0.8) +
  geom_point(data = subset(a375_re, direction == "down"), aes(color = AveExpr), size = 2, alpha = 0.8) +
  geom_smooth(data = subset(a375_re, direction == "unperturbed"), method = "gam", se = FALSE, color = "grey65") +
  geom_smooth(data = subset(a375_re, direction == "up"), method = "gam", se = FALSE, color = "red3") +
  geom_smooth(data = subset(a375_re, direction == "down"), method = "gam", se = FALSE, color = "blue3") +
  geom_smooth(method = "gam", se = FALSE, color = "black", linetype = 2) +
  geom_rug(sides = "bl", alpha = 0.05) +
  scale_color_viridis_c(direction = -1) +
  scale_x_log10() +
  labs(title = "A375 THZ531 vs DMSO 6hr",
       x = "Median Effective Length (bp)",
       y = "logFC") +
  theme_coriell()

p2 <- ggplot(Colo829_re, aes(x = medianEffectiveLength, y = logFC)) +
  geom_point(data = subset(Colo829_re, direction == "unperturbed"), color = "grey80", size = 0.5, alpha = 0.5) +
  geom_point(data = subset(Colo829_re, direction == "up"), aes(color = AveExpr), size = 2, alpha = 0.8) +
  geom_point(data = subset(Colo829_re, direction == "down"), aes(color = AveExpr), size = 2, alpha = 0.8) +
  geom_smooth(data = subset(Colo829_re, direction == "unperturbed"), method = "gam", se = FALSE, color = "grey65") +
  geom_smooth(data = subset(Colo829_re, direction == "up"), method = "gam", se = FALSE, color = "red3") +
  geom_smooth(data = subset(Colo829_re, direction == "down"), method = "gam", se = FALSE, color = "blue3") +
  geom_smooth(method = "gam", se = FALSE, color = "black", linetype = 2) +
  geom_rug(sides = "bl", alpha = 0.05) +
  scale_color_viridis_c(direction = -1) +
  scale_x_log10() +
  labs(title = "Colo829 THZ531 vs DMSO 6hr",
       x = "Median Effective Length (bp)",
       y = "logFC") +
  theme_coriell()

(p1 | p2) +
  plot_annotation(title = "Median Effective Length vs LogFC") & theme_coriell()
ggsave(here("results", "PRJNA766024", "figures", "width_vs_logFC2-re.png"), width = 18, height = 8)
```

### MA plots of REs by effective lengths

```{r}
v1 <- ggplot(a375_re, aes(x = logFC, y = -log10(FDR), color = log10(medianEffectiveLength))) +
  geom_point(alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = 2) +
  scale_color_viridis_c(direction = -1) +
  labs(title = "A375 THZ531 vs DMSO 6hr",
       x = "logFC",
       y = "-log10(FDR)",
       color = "log10(length)") +
  theme_coriell()

v2 <- ggplot(Colo829_re, aes(x = logFC, y = -log10(FDR), color = log10(medianEffectiveLength))) +
  geom_point(alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = 2) +
  scale_color_viridis_c(direction = -1) +
  labs(title = "Colo829 THZ531 vs DMSO 6hr",
       x = "logFC",
       y = "-log10(FDR)",
       color = "log10(length)") +
  theme_coriell()

m1 <- ggplot(a375_re, aes(x = AveExpr, y = logFC, color = log10(medianEffectiveLength))) +
  geom_point(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = 2) +
  scale_color_viridis_c(direction = -1) +
  scale_x_log10() +
  labs(title = "A375 THZ531 vs DMSO 6hr",
       x = "AveExpr",
       y = "logFC",
       color = "log10(length)") +
  theme_coriell()

m2 <- ggplot(Colo829_re, aes(x = AveExpr, y = logFC, color = log10(medianEffectiveLength))) +
  geom_point(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = 2) +
  scale_color_viridis_c(direction = -1) +
  scale_x_log10() +
  labs(title = "Colo829 THZ531 vs DMSO 6hr",
       x = "AveExpr",
       y = "logFC",
       color = "log10(length)") +
  theme_coriell()

(v1 | v2) / (m1 | m2) +
  plot_layout(guides = "collect") & 
  theme_coriell()
ggsave(here("results", "PRJNA766024", "figures", "diffExpr-re-byLength.png"), width = 18, height = 12)
```



