---
title: "PRJNA335388 Analysis"
output:
  html_document:
    df_print: paged
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
source("helpers/helpers.R")
```

## Load Libraries

```{r}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(MultiAssayExperiment))
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(qsmooth))
```

## Read in MultiAssayExperiment Object

```{r}
mae <- readRDS(here("data", "PRJNA335388", "MultiAssayExperiment.rds"))
```

## Read in gene annotations

```{r}
cds <- readRDS(here("doc", "gencode.v26.annotation.cds.rds"))
```

## Extract RE and Gene Counts as matrices

```{r}
genes <- assay(mae, "gene counts")
repeats <- assay(mae, "RE counts")
```

## Restrict REs to five classes

LINE, SINE, long terminal repeats (LTR), SVA, and DNA transposons

```{r}
repeats <- subset(repeats, grepl(rownames(repeats), pattern = "^LINE\\.|^SINE\\.|^LTR\\.|.*\\.SVA\\..*|^DNA\\."))
```

## Restrict genes to protein coding

```{r}
genes <- subset(genes, rownames(genes) %in% cds$gene_name)
```

## Combine both matrices into a single count matrix

```{r}
# check for consistent colnames
all(colnames(genes) == colnames(repeats))

# bind both matrices into single count matrix
counts <- rbind(genes, repeats)
```

## Sum technical replicates across runs and collapse metadata (if present)

Also has the effect of renaming the rows by the BioSample accession.

```{r, warning=FALSE}
counts <- edgeR::sumTechReps(counts, ID = colData(mae)$BioSample)
colData(mae)$bases <- as.integer(colData(mae)$bases)
metadata <- as.data.table(colData(mae))
metadata <- unique(metadata[, .(BioSample, group)])
setDF(metadata, rownames = metadata$BioSample)
```

## DE pipeline

```{r}
design <- model.matrix(~0 + group, data = metadata)
colnames(design) <- gsub(pattern = "^group", replacement = "", x = colnames(design))
y <- DGEList(counts = counts, samples = metadata)
keep <- filterByExpr(y, design = design)
y <- y[keep,, keep.lib.sizes = FALSE]
qs_norm <- qsmooth(y$counts, group_factor = y$samples$group)
disp <- estimateDisp(y = qsmoothData(qs_norm), design = design, robust = TRUE)
fit <- glmQLFit(y = qsmoothData(qs_norm), dispersion = disp$trended.dispersion, design = design, robust = TRUE)
```

## DE testing

```{r}
contrast.matrix <- makeContrasts(
  MDA.Doxorubicin_vs_MDA.DMSO = MDA.doxorubicin_1uM - MDA.dmso,
  MDA.Etoposide_vs_MDA.DMSO = MDA.etoposide_20uM - MDA.dmso,
  MDA.MMS_vs_MDA.DMSO = MDA.mms_40ug_ml - MDA.dmso,
  MDA.Paclitaxel_vs_MDA.DMSO = MDA.paclitaxel_0.2uM - MDA.dmso,
  U2OS.Doxorubicin_vs_U2OS.DMSO = U2OS.doxorubicin_1uM - U2OS.dmso,
  U2OS.Etoposide_vs_U2OS.DMSO = U2OS.etoposide_20uM - U2OS.dmso,
  U2OS.MMS_vs_U2OS.DMSO = U2OS.mms_40ug_ml - U2OS.dmso,
  U2OS.Paclitaxel_vs_U2OS.DMSO = U2OS.paclitaxel_0.2uM - U2OS.dmso,
  
  Doxorubicin_vs_DMSO = ((MDA.doxorubicin_1uM + U2OS.doxorubicin_1uM) / 2) - ((MDA.dmso + U2OS.dmso) / 2),
  Etoposide_vs_DMSO = ((MDA.etoposide_20uM + U2OS.etoposide_20uM) / 2) - ((MDA.dmso + U2OS.dmso) / 2),
  MMS_vs_DMSO = ((MDA.mms_40ug_ml + U2OS.mms_40ug_ml) / 2) - ((MDA.dmso + U2OS.dmso) / 2),
  Paclitaxel_vs_DMSO = ((MDA.paclitaxel_0.2uM + U2OS.paclitaxel_0.2uM) / 2) - ((MDA.dmso + U2OS.dmso) / 2), 
  
  levels = design
)

# create inputs to map over (titles are same as above)
cons <- colnames(contrast.matrix)
names(cons) <- cons

# set global significance levels (used in all downstream plots/analyses)
fc <- 1.5

# map over and extract results
de_results <- lapply(
  cons, 
  get_de_results2, 
  contrast_matrix = contrast.matrix,
  fc = fc,
  glm_fit = fit
)

# extract all results and bind to single df
df <- rbindlist(lapply(de_results, function(x) x[["table"]]), idcol = "contrast")
```

## Calculate normalized counts

```{r}
lcpms <- cpm(qsmoothData(qs_norm), log = TRUE, prior.count = 2)
```

## Save a `SummarizedExperiment` of processed data

Save the:

- Filtered raw counts (gene + RE)
- Normalized counts
- Log counts per million of normalized counts
- Metadata used for DE testing
- Design matrix used for DE testing
- Dispersion estimates for DE testing
- Model fit used in DE testing
- Differential expression results

```{r}
se <- SummarizedExperiment(assays = list("counts" = y$counts,
                                         "qs_norm" = qsmoothData(qs_norm),
                                         "lcpm" = lcpms),
                           colData = metadata,
                           metadata = list("design" = design, 
                                           "contrasts" = contrast.matrix,
                                           "dispersions" = disp,
                                           "fit" = fit,
                                           "de_results" = df))

saveRDS(se, here("results", "PRJNA335388", "rds-files", "se.rds"))
```
