---
title: "PRJNA392914 Analysis"
output:
  html_document:
    df_print: paged
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 4

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
source("helpers/helpers.R")
```

## Load Libraries

```{r}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(MultiAssayExperiment))
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(qsmooth))
```

## Read in MultiAssayExperiment Object

```{r}
mae <- readRDS(here("data", "PRJNA392914", "MultiAssayExperiment.rds"))
```

## Read in gene annotations

```{r}
cds <- readRDS(here("doc", "gencode.v26.annotation.cds.rds"))
```

## Extract RE and Gene Counts as matrices

```{r}
genes <- assay(mae, "gene counts")
repeats <- assay(mae, "RE counts")
```

## Restrict REs to five classes

LINE, SINE, long terminal repeats (LTR), SVA, and DNA transposons

```{r}
repeats <- subset(repeats, grepl(rownames(repeats), pattern = "^LINE\\.|^SINE\\.|^LTR\\.|.*\\.SVA\\..*|^DNA\\."))
```

## Restrict genes to protein coding

```{r}
genes <- subset(genes, rownames(genes) %in% cds$gene_name)
```

## Combine both matrices into a single count matrix

```{r}
# check for consistent colnames
all(colnames(genes) == colnames(repeats))

# bind both matrices into single count matrix
counts <- rbind(genes, repeats)
```

## Sum technical replicates across runs and collapse metadata (if present)

Also has the effect of renaming the rows by the BioSample accession.

```{r, warning=FALSE}
counts <- edgeR::sumTechReps(counts, ID = colData(mae)$BioSample)
colData(mae)$bases <- as.integer(colData(mae)$bases)
metadata <- as.data.table(colData(mae))
metadata <- unique(metadata[, .(BioSample, group)])
setDF(metadata, rownames = metadata$BioSample)
```

## DE pipeline

```{r}
design <- model.matrix(~0 + group, data = metadata)
colnames(design) <- gsub(pattern = "^group", replacement = "", x = colnames(design))
y <- DGEList(counts = counts, samples = metadata)
keep <- filterByExpr(y, design = design)
y <- y[keep,, keep.lib.sizes = FALSE]
qs_norm <- qsmooth(y$counts, group_factor = y$samples$group)
disp <- estimateDisp(y = qsmoothData(qs_norm), design = design, robust = TRUE)
fit <- glmQLFit(y = qsmoothData(qs_norm), dispersion = disp$trended.dispersion, design = design, robust = TRUE)
```

## DE testing

```{r}
contrast.matrix <- makeContrasts(
  PC9_DTP.TSA_vs_PC9_DTP.Untreated = PC9_DTP.TSA.50nM_5hr - PC9_DTP.untreated,
  PC9.Carboplatin_vs_DMSO = PC9.carboplatin.50uM_72hr - PC9.DMSO,
  PC9.Erlotinib_vs_DMSO = PC9.erlotinib.1uM_24hr - PC9.DMSO,
  PC9.TSA_vs_DMSO = PC9.TSA.50nM_5hr - PC9.DMSO,
  PC9_DTP.Untreated_vs_DMSO = PC9_DTP.untreated - PC9.DMSO,
  
  levels = design
)

# create inputs to map over (titles are same as above)
cons <- colnames(contrast.matrix)
names(cons) <- cons

# set global significance levels (used in all downstream plots/analyses)
fc <- 1.5

# map over and extract results
de_results <- lapply(
  cons, 
  get_de_results2, 
  contrast_matrix = contrast.matrix,
  fc = fc,
  glm_fit = fit
)

# extract all results and bind to single df
df <- rbindlist(lapply(de_results, function(x) x[["table"]]), idcol = "contrast")
```

## Write results out

```{r}
# split DE results into gene and RE results and write out
fwrite(
  x = df[feature_id %chin% cds$gene_name],
  file = here("results", "PRJNA392914", "data-files", "dge.tsv"),
  sep = "\t"
  )

fwrite(
  x = df[!feature_id %chin% cds$gene_name],
  file = here("results", "PRJNA392914", "data-files", "dre.tsv"),
  sep = "\t"
)
```

## Calculate normalized counts

```{r}
lcpms <- cpm(qsmoothData(qs_norm), log = TRUE, prior.count = 2)

# split lcpms by gene and by RE
lcpms.gene <- subset(lcpms, rownames(lcpms) %in% cds$gene_name)
lcpms.re <- subset(lcpms, !rownames(lcpms) %in% cds$gene_name)

# Write normalized counts out
saveRDS(lcpms.gene, here("results", "PRJNA392914", "rds-files", "gene-lcpms.rds"))
saveRDS(lcpms.re, here("results", "PRJNA392914", "rds-files", "re-lcpms.rds"))
```

## Save design and contrast matrix

```{r}
saveRDS(design, here("results", "PRJNA392914", "rds-files", "design.rds"))
saveRDS(contrast.matrix, here("results", "PRJNA392914", "rds-files", "contrast-matrix.rds"))
```

---

# OLD CODE

## Create plots for each contrast 

```{r eval=FALSE, include=FALSE}
coi <- names(de_results)
gene_vplots <- vector("list", length(coi))
gene_mdplots <- vector("list", length(coi))
re_vplots <- vector("list", length(coi))
re_mdplots <- vector("list", length(coi))

for (i in seq_along(coi)) {
  gene_vplots[[i]] <- 
    dge %>%
    filter(contrast == coi[[i]]) %>% 
    plot_volcano(fdr = sig, lfc = log2(fc)) +
    ggtitle(titles[[i]])
  
  gene_mdplots[[i]] <- 
    dge %>%
    filter(contrast == coi[[i]]) %>% 
    plot_md(fdr = sig, lfc = log2(fc)) +
    ggtitle(titles[[i]])
  
  re_vplots[[i]] <- 
    dre %>% 
    filter(contrast == coi[[i]]) %>% 
    plot_volcano(fdr = sig, lfc = log2(fc)) +
    ggtitle(titles[[i]])
  
  re_mdplots[[i]] <- 
    dre %>% 
    filter(contrast == coi[[i]]) %>% 
    plot_md(fdr = sig, lfc = log2(fc)) +
    ggtitle(titles[[i]])
}
```

## PCA

```{r eval=FALSE, include=FALSE}
library(FactoMineR)
library(factoextra)


# perform PCA with factominer --------------------------------------------------
pca.gene <- PCA(t(lcpms.gene),
                scale.unit = TRUE,
                graph = FALSE,
                ncp = 5)

pca.re <- PCA(t(lcpms.re),
              scale.unit = TRUE,
              graph = FALSE,
              ncp = 5)

# scree plot -------------------------------------------------------------------
fviz_eig(pca.gene, addlabels = TRUE) +
  ggtitle("Protein Coding Genes") +
  ggsave(here("results", "PRJNA392914", "figures", "gene-scree.png"),
         width = 8,
         height = 5)

fviz_eig(pca.re, addlabels = TRUE) +
  ggtitle("Repetitive Elements") +
  ggsave(here("results", "PRJNA392914", "figures", "re-scree.png"),
         width = 8,
         height = 5)

# extract the individual level coordinates -------------------------------------
ind.gene <- get_pca_ind(pca.gene)
ind.re <- get_pca_ind(pca.re)

# plot -------------------------------------------------------------------------
# genes
ind.gene$coord %>% 
  as_tibble(rownames = "sample_name") %>% 
  left_join(as_tibble(colData(se), rownames = "sample_name"), by = "sample_name") %>% 
  separate(group, into = c("cell_type", "treatment"), sep = "\\.") %>%
  ggplot(aes(x = Dim.1, y = Dim.2, color = treatment, shape = cell_type)) +
  geom_point(size = 5) +
  geom_hline(yintercept = 0, linetype = 3) +
  geom_vline(xintercept = 0, linetype = 3) +
  labs(title = "Protein Coding Genes",
       x = "PC 1",
       y = "PC 2)") +
  scale_color_brewer(palette = "Dark2") +
  theme_light() +
  ggsave(here("results", "PRJNA392914", "figures", "gene-pca.png"),
         width = 8,
         height = 5)
# REs
ind.re$coord %>% 
  as_tibble(rownames = "sample_name") %>% 
  left_join(as_tibble(colData(se), rownames = "sample_name"), by = "sample_name") %>% 
  separate(group, into = c("cell_type", "treatment"), sep = "\\.") %>%
  ggplot(aes(x = Dim.1, y = Dim.2, color = treatment, shape = cell_type)) +
  geom_point(size = 5) +
  geom_hline(yintercept = 0, linetype = 3) +
  geom_vline(xintercept = 0, linetype = 3) +
  labs(title = "Repetitive Elements",
       x = "PC 1",
       y = "PC 2") +
  scale_color_brewer(palette = "Dark2") +
  theme_light() +
  ggsave(here("results", "PRJNA392914", "figures", "re-pca.png"),
         width = 8,
         height = 5)
```

## Heatmaps

```{r eval=FALSE, include=FALSE}
# create column annotation
col_df <- colData(se) %>% 
  as.data.frame() %>% 
  separate(group, into = c("cell_type", "treatment"), sep = "\\.")

# create colors for column annotations
trt_colors <- RColorBrewer::brewer.pal(n = length(unique(col_df$treatment)), "Dark2")
cell_colors <- c("steelblue", "firebrick")
names(cell_colors) <- unique(col_df$cell_type)
names(trt_colors) <- unique(col_df$treatment)

# put colors in list
ann_colors <- list(cell_type = cell_colors, treatment = trt_colors)

# extract only significant genes and REs ---------------------------------------
sig_genes <- df %>% filter(feature_id %in% cds$gene_name & FDR < sig) %>% pull(feature_id)
sig_res <- df %>% filter(!feature_id %in% cds$gene_name & FDR < sig) %>% pull(feature_id)
lcpms.sig_genes <- subset(lcpms.gene, rownames(lcpms.gene) %in% sig_genes)
lcpms.sig_res <- subset(lcpms.re, rownames(lcpms.re) %in% sig_res)

# plot heatmaps ----------------------------------------------------------------
quickmap(lcpms.sig_genes, 
         annotation_col = col_df,
         annotation_colors = ann_colors,
         main = paste("Significant (FDR <", sig, ") Protein Coding Genes"),
         filename = here("results", "PRJNA392914", "figures", "gene-heatmap.png"))

quickmap(lcpms.sig_res, 
         annotation_col = col_df,
         annotation_colors = ann_colors,
         main = paste("Significant (FDR <", sig, ") Repetitive Elements"),
         filename = here("results", "PRJNA392914", "figures", "re-heatmap.png"))
```

## Create patchworks

```{r eval=FALSE, include=FALSE}
library(patchwork)


# Gene-level -------------------------------------------------------------------
(gene_vplots[[1]] + scale_y_continuous(limits = c(0, 15)) + scale_x_continuous(limits = c(-15, 15)) | 
 gene_vplots[[2]] + scale_y_continuous(limits = c(0, 15)) + scale_x_continuous(limits = c(-15, 15)) | 
 gene_vplots[[3]] + scale_y_continuous(limits = c(0, 15)) + scale_x_continuous(limits = c(-15, 15))) / 
(gene_vplots[[4]] + scale_y_continuous(limits = c(0, 15)) + scale_x_continuous(limits = c(-15, 15)) |
 gene_vplots[[5]] + scale_y_continuous(limits = c(0, 15)) + scale_x_continuous(limits = c(-15, 15)) |
 plot_spacer()) +
  plot_layout(guides = "collect") +
  plot_annotation("Protein Coding Gene Expression") +
  ggsave(here("results", "PRJNA392914", "figures", "gene-vplots.png"),
         width = 18, 
         height = 10)

(gene_mdplots[[1]] + scale_y_continuous(limits = c(-15, 15)) | 
 gene_mdplots[[2]] + scale_y_continuous(limits = c(-15, 15)) | 
 gene_mdplots[[3]] + scale_y_continuous(limits = c(-15, 15))) / 
(gene_mdplots[[4]] + scale_y_continuous(limits = c(-15, 15)) |
 gene_mdplots[[5]] + scale_y_continuous(limits = c(-15, 15)) |
 plot_spacer()) +
  plot_layout(guides = "collect") +
  plot_annotation("Protein Coding Gene Expression") +
  ggsave(here("results", "PRJNA392914", "figures", "gene-mdplots.png"),
         width = 18, 
         height = 10)

# RE-level ---------------------------------------------------------------------
(re_vplots[[1]] + scale_y_continuous(limits = c(0, 12)) + scale_x_continuous(limits = c(-10, 10)) | 
 re_vplots[[2]] + scale_y_continuous(limits = c(0, 12)) + scale_x_continuous(limits = c(-10, 10)) | 
 re_vplots[[3]] + scale_y_continuous(limits = c(0, 12)) + scale_x_continuous(limits = c(-10, 10))) / 
(re_vplots[[4]] + scale_y_continuous(limits = c(0, 12)) + scale_x_continuous(limits = c(-10, 10)) |
 re_vplots[[5]] + scale_y_continuous(limits = c(0, 12)) + scale_x_continuous(limits = c(-10, 10)) |
 plot_spacer()) +
  plot_layout(guides = "collect") +
  plot_annotation("RE Expression") +
  ggsave(here("results", "PRJNA392914", "figures", "re-vplots.png"),
         width = 18, 
         height = 10)

(re_mdplots[[1]] + scale_y_continuous(limits = c(-10, 10)) | 
 re_mdplots[[2]] + scale_y_continuous(limits = c(-10, 10)) | 
 re_mdplots[[3]] + scale_y_continuous(limits = c(-10, 10))) / 
(re_mdplots[[4]] + scale_y_continuous(limits = c(-10, 10)) |
 re_mdplots[[5]] + scale_y_continuous(limits = c(-10, 10)) |
 plot_spacer()) +
  plot_layout(guides = "collect") +
  plot_annotation("RE Expression") +
  ggsave(here("results", "PRJNA392914", "figures", "re-mdplots.png"),
         width = 18, 
         height = 10)
```

## Create Family-Level Dotplots for significant REs

```{r eval=FALSE, include=FALSE}
dotplots <- vector("list", length(names(de_results)))
for (i in seq_along(names(de_results))) {
  dotplots[[i]] <- plot_fam_counts(dre, names(de_results)[[i]], n_dots = 20)
}

(dotplots[[1]] + ggtitle(titles[[1]]) |
 dotplots[[2]] + ggtitle(titles[[2]]) |
 dotplots[[3]] + ggtitle(titles[[3]])) /
(dotplots[[4]] + ggtitle(titles[[4]]) |
 dotplots[[5]] + ggtitle(titles[[5]]) |
 plot_spacer()) +
  plot_annotation("Family-level Expression of Significant REs") +
  ggsave(here("results", "PRJNA392914", "figures", "dotplots.png"),
        width = 18,
        height = 10)
```
