---
title: "PRJNA670901 Analysis"
output:
  html_document:
    df_print: paged
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
source("helpers/helpers.R")
```

## Load Libraries

```{r}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(MultiAssayExperiment))
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(qsmooth))
suppressPackageStartupMessages(library(quantro))
suppressPackageStartupMessages(library(doParallel))
registerDoParallel(cores = 12)
```

## Create results directory for project if one doesn't exist

```{r}
if (!dir.exists(here("results", "PRJNA670901"))) {
  dir.create(here("results", "PRJNA670901", "rds-files"), recursive = TRUE)
  dir.create(here("results", "PRJNA670901", "data-files"))
}
```

## Read in MultiAssayExperiment Object

```{r}
mae <- readRDS(here("data", "PRJNA670901", "MultiAssayExperiment.rds"))
```

## Read in gene annotations

```{r}
cds <- readRDS(here("doc", "gencode.v26.annotation.cds.rds"))
```

## Extract RE and Gene Counts as matrices

```{r}
genes <- assay(mae, "gene counts")
repeats <- assay(mae, "RE counts")
```

## Restrict REs to five classes

LINE, SINE, long terminal repeats (LTR), SVA, and DNA transposons

```{r}
repeats <- subset(repeats, grepl(rownames(repeats), pattern = "^LINE\\.|^SINE\\.|^LTR\\.|.*\\.SVA\\..*|^DNA\\."))
```

## Restrict genes to protein coding

```{r}
genes <- subset(genes, rownames(genes) %in% cds$gene_name)
```

## Combine both matrices into a single count matrix

```{r}
# check for consistent colnames
all(colnames(genes) == colnames(repeats))

# bind both matrices into single count matrix
counts <- rbind(genes, repeats)
```

## Sum technical replicates across runs and collapse metadata (if present)

Also has the effect of renaming the rows by the BioSample accession.

```{r}
counts <- sumTechReps(counts, ID = colData(mae)$BioSample)
colData(mae)$bases <- as.integer(colData(mae)$bases)
metadata <- as.data.table(colData(mae))
metadata <- unique(metadata[, .(BioSample, group)])
setDF(metadata, rownames = metadata$BioSample)

# Add additional metadata columns testing
metadata$cell_line <- tstrsplit(metadata$group, "\\.")[[1]]
metadata$treatment <- tstrsplit(metadata$group, "\\.")[[2]]
```

## Check for global differences between groups with `quantro`

```{r}
qtest.cell_line <- quantro(counts, groupFactor = metadata$cell_line, B = 1e3)
perm_pval.cell_line <- quantroPvalPerm(qtest.cell_line)
perm_pval <- perm_pval.cell_line
```

## DE pipeline

```{r}
# Define experimental design
design <- model.matrix(~0 + group, data = metadata)
colnames(design) <- gsub(pattern = "^group", replacement = "", x = colnames(design))

# Define contrasts to test
contrast.matrix <- makeContrasts(
  HH.Azacitidine.125nM_vs_HH.Untreated = HH.azacitidine_125nM - HH.untreated,
  HH.Azacitidine.250nM_vs_HH.Untreated = HH.azacitidine_250nM - HH.untreated,
  HH.Azacitidine.50nM_vs_HH.Untreated = HH.azacitidine_50nM - HH.untreated,
  HH.Decitabine.12.5nM_HH.Untreated = HH.dac_12.5nM - HH.untreated,
  HH.Decitabine.25nM_vs_HH.Untreated = HH.dac_25nM - HH.untreated,
  HH.Decitabine.5nM_vs_HH.Untreated = HH.dac_5nM - HH.untreated,
  PF382.Azacitidine.125nM_vs_PF382.Untreated = PF382.azacitidine_125nM - PF382.untreated,
  PF382.Azacitidine.250nM_vs_PF382.Untreated = PF382.azacitidine_250nM - PF382.untreated,
  PF382.Azacitidine.500nM_vs_PF382.Untreated = PF382.azacitidine_500nM - PF382.untreated,
  PF382.Decitabine.12.5nM_vs_PF382.Untreated = PF382.dac_12.5nM - PF382.untreated,
  PF382.Decitabine.25nM_vs_PF382.Untreated = PF382.dac_25nM - PF382.untreated,
  PF382.Decitabine.5nM_vs_PF382.Untreated = PF382.dac_5nM - PF382.untreated,
  TLOm1.Azacitidine.125nM_vs_TLOm1.Untreated = TLOm1.azacitidine_125nM - TLOm1.untreated,
  TLOm1.Azacitidine.250nM_vs_TLOm1.Untreated = TLOm1.azacitidine_250nM - TLOm1.untreated,
  TLOm1.Azacitidine.500nM_vs_TLOm1.Untreated = TLOm1.azacitidine_500nM - TLOm1.untreated,
  TLOm1.Decitabine.12.5nM_vs_TLOm1.Untreated = TLOm1.dac_12.5nM - TLOm1.untreated,
  TLOm1.Decitabine.25nM_vs_TLOm1.Untreated = TLOm1.dac_25nM - TLOm1.untreated,
  TLOm1.Decitabine.5nM_vs_TLOm1.Untreated = TLOm1.dac_5nM - TLOm1.untreated,
  
  HH.Azacitidine_vs_HH.untreated = ((HH.azacitidine_125nM + HH.azacitidine_250nM + HH.azacitidine_50nM) / 3) - HH.untreated,
  HH.Decitabine_vs_HH.untreated = ((HH.dac_12.5nM + HH.dac_25nM + HH.dac_5nM) / 3) - HH.untreated,
  PF382.azacitidine = ((PF382.azacitidine_125nM + PF382.azacitidine_250nM + PF382.azacitidine_500nM) / 3) - PF382.untreated,
  PF382.Decitabine_vs_PF382.Untreated = ((PF382.dac_12.5nM + PF382.dac_25nM + PF382.dac_5nM) / 3) - PF382.untreated,
  TLOm1.Azacitidine_vs_TLOm1.Untreated = ((TLOm1.azacitidine_125nM + TLOm1.azacitidine_250nM + TLOm1.azacitidine_500nM) / 3) - TLOm1.untreated,
  TLOm1.Decitabine_vs_TLOm1.Untreated = ((TLOm1.dac_12.5nM + TLOm1.dac_25nM + TLOm1.dac_5nM) / 3) - TLOm1.untreated,
  
  Azacitidine.125nM_vs_Untreated = ((HH.azacitidine_125nM + PF382.azacitidine_125nM + TLOm1.azacitidine_125nM) / 3) - ((HH.untreated + PF382.untreated + TLOm1.untreated) / 3),
  Azacitidine.250nM_vs_Untreated = ((HH.azacitidine_250nM + PF382.azacitidine_250nM + TLOm1.azacitidine_250nM) / 3) - ((HH.untreated + PF382.untreated + TLOm1.untreated) / 3),
  Azacitidine.500nM_vs_Untreated = ((HH.azacitidine_50nM + PF382.azacitidine_500nM + TLOm1.azacitidine_500nM) / 3) -  ((HH.untreated + PF382.untreated + TLOm1.untreated) / 3),
  
  Decitabine.12.5nM_vs_Untreated = ((HH.dac_12.5nM + PF382.dac_12.5nM + TLOm1.dac_12.5nM) / 3) - ((HH.untreated + PF382.untreated + TLOm1.untreated) / 3),
  Decitabine.25nM_vs_Untreated = ((HH.dac_25nM + PF382.dac_25nM + TLOm1.dac_25nM) / 3) - ((HH.untreated + PF382.untreated + TLOm1.untreated) / 3),
  Decitabine.5nM_vs_Untreated = ((HH.dac_5nM + PF382.dac_5nM + TLOm1.dac_5nM) / 3) - ((HH.untreated + PF382.untreated + TLOm1.untreated) / 3),
  
  levels = design
)

# Create DGEList object from counts and filter for expression
y <- DGEList(counts = counts, samples = metadata)
keep <- filterByExpr(y, design = design)
y <- y[keep,, keep.lib.sizes = FALSE]

# If global differences exist use QSmooth normalization, if not use TMM
if (perm_pval < 0.01) {
  norm_method <- "qs"
  qs <- qsmooth(y$counts, group_factor = y$samples$group)
  fit <- lmFit(log2(qsmoothData(qs) + 2), design)
  fit <- contrasts.fit(fit, contrast.matrix)
} else {
  norm_method <- "tmm"
  y <- calcNormFactors(y, method = "TMM")
  y <- estimateDisp(y, design, robust = TRUE)
  fit <- glmQLFit(y, design, robust = TRUE)
  plotQLDisp(fit)
}
```

## Test for differential expression above a fold-change

Use either `treat` or `glmTreat` depending on the fit type to test for 
differential expression above a fold-change.

```{r}
# Define fold-change that features must exceed to be considered for significance
FC <- 1.5

contrast_names <- colnames(contrast.matrix)
if (norm_method == "qs") {
  fit2 <- treat(fit, lfc = log2(FC), trend = TRUE, robust = TRUE)
  plotSA(fit2)
  res_dts <- get_treat(contrast_names, fit2)
} else {
  res_dts <- get_glmTreat(contrast_names, fit, contrast.matrix, FC)
  }

# extract all results and bind to single df
df <- rbindlist(res_dts, idcol = "contrast")
```

## Save a `SummarizedExperiment` of processed data

```{r}
# Calculate logcounts of filtered, normalized data
if (norm_method == "qs") {
  logcounts <- log2(qsmoothData(qs) + 2)
} else {
  logcounts <- cpm(y, log = TRUE, prior.count = 2)
}

# Export results as SummarizedExperiment
se <- SummarizedExperiment(
  assays = list("counts" = y$counts, "logcounts" = logcounts),
  colData = metadata,
  metadata = list("fit" = fit, "de" = df)
  )
saveRDS(se, here("results", "PRJNA670901", "rds-files", "se.rds"))
```
