---
title: "PRJNA260609 Analysis"
output:
  html_document:
    df_print: paged
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 4
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
source("helpers/helpers.R")
```

## Read in MultiAssayExperiment Object

```{r}
library(here)
library(data.table)
library(MultiAssayExperiment)


mae <- readRDS(here("data", "PRJNA260609", "MultiAssayExperiment.rds"))
```

## Read in gene annotations

```{r}
cds <- readRDS(here("doc", "gencode.v26.annotation.cds.rds"))
```

## Extract RE and Gene Counts as matrices

```{r}
genes <- assay(mae, "gene counts")
repeats <- assay(mae, "RE counts")
```

## Restrict REs to five classes

LINE, SINE, long terminal repeats (LTR), SVA, and DNA transposons

```{r}
repeats <- subset(repeats, grepl(rownames(repeats), pattern = "^LINE\\.|^SINE\\.|^LTR\\.|.*\\.SVA\\..*|^DNA\\."))
```

## Restrict genes to protein coding

```{r}
genes <- subset(genes, rownames(genes) %in% cds$gene_name)
```

## Combine both matrices into a single count matrix

```{r}
# check for consistent colnames
all(colnames(genes) == colnames(repeats))

# bind both matrices into single count matrix
counts <- rbind(genes, repeats)
```

## DE pipeline

```{r}
colData(mae)$mutation <- factor(rep(c("siSMYD2", "siControl"), each = 3), levels = c("siControl", "siSMYD2"))
colData(mae)$drug <- factor(rep(c("Rapamycin", "BIX01294", "Control"), 2), levels = c("Control", "BIX01294", "Rapamycin"))
design <- model.matrix(~mutation + drug, data = colData(mae))
y <- DGEList(counts = counts,
             samples = colData(mae))
keep <- filterByExpr(y, design = design)
y <- y[keep,, keep.lib.sizes = FALSE]
y <- calcNormFactors(y, design = design, method = "TMM")
y <- estimateDisp(y, design = design, robust = TRUE)
fit <- glmQLFit(y, design = design, robust = TRUE)
```

## DE testing

```{r}
bix.res <- setDT(coriell::edger_to_df(glmTreat(fit, coef = 3, lfc = log2(1.5))))
bix.res[, contrast := "BIX01294_vs_Control"]
bix.res <- dplyr::select(bix.res, contrast, dplyr::everything())

rapa.res <- setDT(coriell::edger_to_df(glmTreat(fit, coef = 4, lfc = log2(1.5))))
rapa.res[, contrast := "Rapamycin_vs_Control"]
rapa.res <- dplyr::select(rapa.res, contrast, dplyr::everything())

df <- rbindlist(list(bix.res, rapa.res))
```

## Write results out

```{r}
# split DE results into gene and RE results and write out
fwrite(
  x = df[feature_id %chin% cds$gene_name],
  file = here("results", "PRJNA260609", "data-files", "dge.tsv"),
  sep = "\t"
  )

fwrite(
  x = df[!feature_id %chin% cds$gene_name],
  file = here("results", "PRJNA260609", "data-files", "dre.tsv"),
  sep = "\t"
)
```

## Calculate normalized counts

```{r}
lcpms <- cpm(y, log = TRUE, prior.count = 2)

# split lcpms by gene and by RE
lcpms.gene <- subset(lcpms, rownames(lcpms) %in% cds$gene_name)
lcpms.re <- subset(lcpms, !rownames(lcpms) %in% cds$gene_name)

# Write normalized counts out
saveRDS(lcpms.gene, here("results", "PRJNA260609", "rds-files", "gene-lcpms.rds"))
saveRDS(lcpms.re, here("results", "PRJNA260609", "rds-files", "re-lcpms.rds"))
```

## Save design and contrast matrix

```{r}
saveRDS(design, here("results", "PRJNA260609", "rds-files", "design.rds"))
#saveRDS(contrast.matrix, here("results", "PRJNA260609", "rds-files", "contrast-matrix.rds"))
```
