---
title: "PRJNA515531 Analysis"
output:
  html_document:
    df_print: paged
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
source("helpers/helpers.R")
```

## Load Libraries

```{r}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(MultiAssayExperiment))
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(qsmooth))
suppressPackageStartupMessages(library(quantro))
suppressPackageStartupMessages(library(doParallel))
registerDoParallel(cores = 12)
```

## Create results directory for project if one doesn't exist

```{r}
if (!dir.exists(here("results", "PRJNA515531"))) {
  dir.create(here("results", "PRJNA515531", "rds-files"), recursive = TRUE)
  dir.create(here("results", "PRJNA515531", "data-files"))
}
```

## Read in MultiAssayExperiment Object

```{r}
mae <- readRDS(here("data", "PRJNA515531", "MultiAssayExperiment.rds"))
```

## Read in gene annotations

```{r}
cds <- readRDS(here("doc", "gencode.v26.annotation.cds.rds"))
```

## Extract RE and Gene Counts as matrices

```{r}
genes <- assay(mae, "gene counts")
repeats <- assay(mae, "RE counts")
```

## Restrict REs to five classes

LINE, SINE, long terminal repeats (LTR), SVA, and DNA transposons

```{r}
repeats <- subset(repeats, grepl(rownames(repeats), pattern = "^LINE\\.|^SINE\\.|^LTR\\.|.*\\.SVA\\..*|^DNA\\."))
```

## Restrict genes to protein coding

```{r}
genes <- subset(genes, rownames(genes) %in% cds$gene_name)
```

## Combine both matrices into a single count matrix

```{r}
# check for consistent colnames
all(colnames(genes) == colnames(repeats))

# bind both matrices into single count matrix
counts <- rbind(genes, repeats)
```

## Sum technical replicates across runs and collapse metadata (if present)

Also has the effect of renaming the rows by the BioSample accession.

```{r}
counts <- sumTechReps(counts, ID = colData(mae)$BioSample)
colData(mae)$bases <- as.integer(colData(mae)$bases)
metadata <- as.data.table(colData(mae))
metadata <- unique(metadata[, .(BioSample, group)])
setDF(metadata, rownames = metadata$BioSample)
```

## Check for global differences between groups with `quantro`

```{r}
qtest <- quantro(counts, groupFactor = metadata$group, B = 1e3)
perm_pval <- quantroPvalPerm(qtest)
```

## DE pipeline

```{r}
# Define experimental design
design <- model.matrix(~0 + group, data = metadata)
colnames(design) <- gsub(pattern = "^group", replacement = "", x = colnames(design))

# Define contrasts to test
contrast.matrix <- makeContrasts(
  BT549.Abemaciclib_vs_BT549.DMSO = BT549.Abemaciclib - BT549.DMSO,
  BT549.Alvocidib_vs_BT549.DMSO = BT549.Alvocidib - BT549.DMSO,
  BT549.Palbociclib_vs_BT549.DMSO = BT549.Palbociclib - BT549.DMSO,
  BT549.Ribociclib_vs_BT549.DMSO = BT549.Ribociclib - BT549.DMSO,
  
  HCC1806.Abemaciclib_vs_HCC1806.DMSO = HCC1806.Abemaciclib - HCC1806.DMSO,
  HCC1806.Alvocidib_vs_HCC1806.DMSO = HCC1806.Alvocidib - HCC1806.DMSO,
  HCC1806.Palbociclib_vs_HCC1806.DMSO = HCC1806.Palbociclib - HCC1806.DMSO,
  HCC1806.Ribociclib_vs_HCC1806.DMSO = HCC1806.Ribociclib - HCC1806.DMSO,
  
  Hs578T.Abemaciclib_vs_Hs578T.DMSO = Hs578T.Abemaciclib - Hs578T.DMSO,
  Hs578T.Alvocidib_vs_Hs578T.DMSO = Hs578T.Alvocidib - Hs578T.DMSO,
  Hs578T.Palbociclib_vs_Hs578T.DMSO = Hs578T.Palbociclib - Hs578T.DMSO,
  Hs578T.Ribociclib_vs_Hs578T.DMSO = Hs578T.Ribociclib - Hs578T.DMSO,
  
  MCF7.Abemaciclib_vs_MCF7.DMSO = MCF7.Abemaciclib - MCF7.DMSO,
  MCF7.Alvocidib_vs_MCF7.DMSO = MCF7.Alvocidib - MCF7.DMSO,
  MCF7.Palbociclib_vs_MCF7.DMSO = MCF7.Palbociclib - MCF7.DMSO,
  MCF7.Ribociclib_vs_MCF7.DMSO = MCF7.Ribociclib - MCF7.DMSO,
  
  PDX1258.Abemaciclib_vs_PDX1258.DMSO = PDX1258.Abemaciclib - PDX1258.DMSO,
  PDX1258.Alvocidib_vs_PDX1258.DMSO = PDX1258.Alvocidib - PDX1258.DMSO,
  PDX1258.Palbociclib_vs_PDX1258.DMSO = PDX1258.Palbociclib - PDX1258.DMSO,
  PDX1258.Ribociclib_vs_PDX1258.DMSO = PDX1258.Ribociclib - PDX1258.DMSO,
  
  PDXHCI002.Abemaciclib_vs_PDXHCI002.DMSO = PDXHCI002.Abemaciclib - PDXHCI002.DMSO,
  PDXHCI002.Alvocidib_vs_PDXHCI002.DMSO = PDXHCI002.Alvocidib - PDXHCI002.DMSO,
  PDXHCI002.Palbociclib_vs_PDXHCI002.DMSO = PDXHCI002.Palbociclib - PDXHCI002.DMSO,
  PDXHCI002.Ribociclib_vs_PDXHCI002.DMSO = PDXHCI002.Ribociclib - PDXHCI002.DMSO,
  
  T47D.Abemaciclib_vs_T47D.DMSO = T47D.Abemaciclib - T47D.DMSO,
  T47D.Alvocidib_vs_T47D.DMSO = T47D.Alvocidib - T47D.DMSO,
  T47D.Palbociclib_vs_T47D.DMSO = T47D.Palbociclib - T47D.DMSO,
  T47D.Ribociclib_vs_T47D.DMSO = T47D.Ribociclib - T47D.DMSO,
  
  levels = design
)

# Create DGEList object from counts and filter for expression
y <- DGEList(counts = counts, samples = metadata)
keep <- filterByExpr(y, design = design)
y <- y[keep,, keep.lib.sizes = FALSE]

# If global differences exist use QSmooth normalization, if not use TMM
if (perm_pval < 0.01) {
  norm_method <- "qs"
  qs <- qsmooth(y$counts, group_factor = y$samples$group)
  fit <- lmFit(log2(qsmoothData(qs) + 2), design)
  fit <- contrasts.fit(fit, contrast.matrix)
} else {
  norm_method <- "tmm"
  y <- calcNormFactors(y, method = "TMM")
  y <- estimateDisp(y, design, robust = TRUE)
  fit <- glmQLFit(y, design, robust = TRUE)
  plotQLDisp(fit)
}
```

## Test for differential expression above a fold-change

Use either `treat` or `glmTreat` depending on the fit type to test for 
differential expression above a fold-change.

```{r}
# Define fold-change that features must exceed to be considered for significance
FC <- 1.5

contrast_names <- colnames(contrast.matrix)
if (norm_method == "qs") {
  fit2 <- treat(fit, lfc = log2(FC), trend = TRUE, robust = TRUE)
  plotSA(fit2)
  res_dts <- get_treat(contrast_names, fit2)
} else {
  res_dts <- get_glmTreat(contrast_names, fit, contrast.matrix, FC)
  }

# extract all results and bind to single df
df <- rbindlist(res_dts, idcol = "contrast")
```

## Save a `SummarizedExperiment` of processed data

Save the:

- Filtered raw counts (gene + RE)
- Metadata used for DE testing
- Design matrix used for DE testing
- Model fit used in DE testing
- Differential expression results

```{r}
se <- SummarizedExperiment(assays = list("counts" = y$counts),
                           colData = metadata,
                           metadata = list("design" = design, 
                                           "contrasts" = contrast.matrix,
                                           "fit" = fit,
                                           "de" = df))

saveRDS(se, here("results", "PRJNA515531", "rds-files", "se.rds"))
```
