---
title: "PRJNA471642 Analysis"
output:
  html_document:
    df_print: paged
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 4
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

source("helpers/helpers.R")
```

## Read in SummarizedExperiment Object

```{r}
library(here)
library(SummarizedExperiment)


se <- readRDS(here("data", "PRJNA471642", "summarized_experiment.rds"))
```

## Read in gene annotations

```{r}
cds <- read_rds(here("doc", "gencode.v26.annotation.cds.rds"))
```

## Filter out RNA REs from counts assay

```{r}
se <- subset(se, !str_detect(rownames(se), "tRNA|snRNA|scRNA|rRNA|srpRNA|Simple_repeat|Low_complexity"))
```

## DE pipeline

This DE analysis is modified because the study design is a time-series

```{r}
library(edgeR)
library(coriell)


Cell_line <- factor(colData(se)$cell_line)
Time <- factor(colData(se)$time_hr)
design <- model.matrix(~Cell_line + Time)

y <- DGEList(counts = assay(se, "counts"),
             samples = colData(se))
keep <- filterByExpr(y, design = design)
y <- y[keep,, keep.lib.sizes = FALSE]
y <- calcNormFactors(y, design = design, method = "TMM")
y <- estimateDisp(y, design = design, robust = TRUE)
fit <- glmQLFit(y, design = design, robust = TRUE)
```

## DE testing

Differential expression testing is set up differently for this dataset

```{r}
# set global significance levels
fc <- 1.5
sig <- 0.1

# perfom DE testing of each coefficient
mec1 <- glmTreat(fit, coef = 2, lfc = log2(fc)) %>% edger_to_df()
time_2hr <- glmTreat(fit, coef = 3, lfc = log2(fc)) %>% edger_to_df()
time_6hr <- glmTreat(fit, coef = 4, lfc = log2(fc)) %>% edger_to_df()
time_24hr <- glmTreat(fit, coef = 5, lfc = log2(fc)) %>% edger_to_df()

# bind results into single df
results_list <- list("MEC1_vs_GM12878" = mec1,
                     "SD6.2hr_vs_SD6.0hr" = time_2hr,
                     "SD6.6hr_vs_SD6.0hr" = time_6hr,
                     "SD6.24hr_vs_SD6.0hr" = time_24hr)

df <- bind_rows(results_list, .id = "contrast")
```

## Write results out

```{r}
# split DE results into gene and RE results and write out
dge <- df %>% 
  filter(feature_id %in% cds$gene_name) %>% 
  write_tsv(here("results", "PRJNA471642", "data-files", "dge.tsv"))

dre <- df %>% 
  filter(!feature_id %in% cds$gene_name) %>% 
  write_tsv(here("results", "PRJNA471642", "data-files", "dre.tsv"))
```

## Calculate normalized counts

```{r}
lcpms <- cpm(y, log = TRUE, prior.count = 2)

# split lcpms by gene and by RE
lcpms.gene <- subset(lcpms, rownames(lcpms) %in% cds$gene_name)
lcpms.re <- subset(lcpms, !rownames(lcpms) %in% cds$gene_name)

# Write normalized counts out
write_rds(lcpms.gene, here("results", "PRJNA471642", "rds-files", "gene-lcpms.rds"))
write_rds(lcpms.re, here("results", "PRJNA471642", "rds-files", "re-lcpms.rds"))
```

---

# OLD CODE

## Create plots for each contrast 

```{r eval=FALSE, include=FALSE}
library(patchwork)


# Gene level plots
v1.gene <- plot_volcano({time_2hr %>% 
                         filter(feature_id %in% cds$gene_name)}, 
                        lfc = log2(fc), 
                        fdr = sig) + ggtitle("SD6.125nm 2hr vs Baseline 0hr")

v2.gene <- plot_volcano({time_6hr %>% 
                         filter(feature_id %in% cds$gene_name)}, 
                        lfc = log2(fc), 
                        fdr = sig) + ggtitle("SD6.125nm 6hr vs Baseline 0hr")

v3.gene <- plot_volcano({time_24hr %>% 
                         filter(feature_id %in% cds$gene_name)}, 
                        lfc = log2(fc), 
                        fdr = sig) + ggtitle("SD6.125nm 24hr vs Baseline 0hr")

m1.gene <- plot_md({time_2hr %>% 
                         filter(feature_id %in% cds$gene_name)}, 
                        lfc = log2(fc), 
                        fdr = sig) + ggtitle("SD6.125nm 2hr vs Baseline 0hr")

m2.gene <- plot_md({time_6hr %>% 
                         filter(feature_id %in% cds$gene_name)}, 
                        lfc = log2(fc), 
                        fdr = sig) + ggtitle("SD6.125nm 6hr vs Baseline 0hr")

m3.gene <- plot_md({time_24hr %>% 
                         filter(feature_id %in% cds$gene_name)}, 
                        lfc = log2(fc), 
                        fdr = sig) + ggtitle("SD6.125nm 24hr vs Baseline 0hr")

# RE level
v1.re <- plot_volcano({time_2hr %>% 
                         filter(!feature_id %in% cds$gene_name)}, 
                        lfc = log2(fc), 
                        fdr = sig) + ggtitle("SD6.125nm 2hr vs Baseline 0hr")

v2.re <- plot_volcano({time_6hr %>% 
                         filter(!feature_id %in% cds$gene_name)}, 
                        lfc = log2(fc), 
                        fdr = sig) + ggtitle("SD6.125nm 6hr vs Baseline 0hr")

v3.re <- plot_volcano({time_24hr %>% 
                         filter(!feature_id %in% cds$gene_name)}, 
                        lfc = log2(fc), 
                        fdr = sig) + ggtitle("SD6.125nm 24hr vs Baseline 0hr")

m1.re <- plot_md({time_2hr %>% 
                         filter(!feature_id %in% cds$gene_name)}, 
                        lfc = log2(fc), 
                        fdr = sig) + ggtitle("SD6.125nm 2hr vs Baseline 0hr")

m2.re <- plot_md({time_6hr %>% 
                         filter(!feature_id %in% cds$gene_name)}, 
                        lfc = log2(fc), 
                        fdr = sig) + ggtitle("SD6.125nm 6hr vs Baseline 0hr")

m3.re <- plot_md({time_24hr %>% 
                         filter(!feature_id %in% cds$gene_name)}, 
                        lfc = log2(fc), 
                        fdr = sig) + ggtitle("SD6.125nm 24hr vs Baseline 0hr")

# Create patchworks
(v1.gene + scale_y_continuous(limits = c(0, 4)) + scale_x_continuous(limits = c(-10, 10)) |
 v2.gene + scale_y_continuous(limits = c(0, 4)) + scale_x_continuous(limits = c(-10, 10)) | 
 v3.gene + scale_y_continuous(limits = c(0, 4)) + scale_x_continuous(limits = c(-10, 10))) / 
(m1.gene + scale_y_continuous(limits = c(-10, 10)) | 
 m2.gene + scale_y_continuous(limits = c(-10, 10)) | 
 m3.gene + scale_y_continuous(limits = c(-10, 10))) +
  plot_annotation("Protein Coding Gene Expression") +
  ggsave(here("results", "PRJNA471642", "figures", "gene-patchwork.png"),
         width = 18,
         height = 10)

(v1.re + scale_y_continuous(limits = c(0, 4)) + scale_x_continuous(limits = c(-8, 8)) |
 v2.re + scale_y_continuous(limits = c(0, 4)) + scale_x_continuous(limits = c(-8, 8)) | 
 v3.re + scale_y_continuous(limits = c(0, 4)) + scale_x_continuous(limits = c(-8, 8))) / 
(m1.re + scale_y_continuous(limits = c(-8, 8))  | 
 m2.re + scale_y_continuous(limits = c(-8, 8))  | 
 m3.re + scale_y_continuous(limits = c(-8, 8)) ) +
  plot_annotation("RE Expression") +
  ggsave(here("results", "PRJNA471642", "figures", "re-patchwork.png"),
         width = 18,
         height = 10)
```

## PCA

```{r eval=FALSE, include=FALSE}
library(FactoMineR)
library(factoextra)


# perform PCA with factominer --------------------------------------------------
pca.gene <- PCA(t(lcpms.gene),
                scale.unit = TRUE,
                graph = FALSE,
                ncp = 5)

pca.re <- PCA(t(lcpms.re),
              scale.unit = TRUE,
              graph = FALSE,
              ncp = 5)

# scree plot -------------------------------------------------------------------
fviz_eig(pca.gene, addlabels = TRUE) +
  ggtitle("Protein Coding Genes") +
  ggsave(here("results", "PRJNA471642", "figures", "gene-scree.png"),
         width = 8,
         height = 5)

fviz_eig(pca.re, addlabels = TRUE) +
  ggtitle("Repetitive Elements") +
  ggsave(here("results", "PRJNA471642", "figures", "re-scree.png"),
         width = 8,
         height = 5)

# extract the individual level coordinates -------------------------------------
ind.gene <- get_pca_ind(pca.gene)
ind.re <- get_pca_ind(pca.re)

# plot -------------------------------------------------------------------------
# genes
ind.gene$coord %>% 
  as_tibble(rownames = "sample_name") %>% 
  left_join(as_tibble(colData(se), rownames = "sample_name"), by = "sample_name") %>% 
  mutate(time_hr = factor(time_hr)) %>% 
  ggplot(aes(x = Dim.1, y = Dim.2, color = cell_line, shape = time_hr)) +
  geom_point(size = 5) +
  geom_hline(yintercept = 0, linetype = 3) +
  geom_vline(xintercept = 0, linetype = 3) +
  labs(title = "Protein Coding Genes",
       x = "PC 1",
       y = "PC 2") +
  scale_color_brewer(palette = "Dark2") +
  theme_light() +
  ggsave(here("results", "PRJNA471642", "figures", "gene-pca.png"),
         width = 8,
         height = 5)
# REs
ind.re$coord %>% 
  as_tibble(rownames = "sample_name") %>% 
  left_join(as_tibble(colData(se), rownames = "sample_name"), by = "sample_name") %>% 
  mutate(time_hr = factor(time_hr)) %>% 
  ggplot(aes(x = Dim.1, y = Dim.2, color = cell_line, shape = time_hr)) +
  geom_point(size = 5) +
  geom_hline(yintercept = 0, linetype = 3) +
  geom_vline(xintercept = 0, linetype = 3) +
  labs(title = "Repetitive Elements",
       x = "PC 1",
       y = "PC 2") +
  scale_color_brewer(palette = "Dark2") +
  theme_light() +
  ggsave(here("results", "PRJNA471642", "figures", "re-pca.png"),
         width = 8,
         height = 5)
```

## Heatmaps

```{r eval=FALSE, include=FALSE}
# create column annotation
col_df <- colData(se)[, c("cell_line", "time_hr")] %>% 
  as.data.frame()

# create colors for column annotations
cell_colors <- c("MEC1" = "#1B9E77", "GM12878" = "#D95F02")
time_colors <- RColorBrewer::brewer.pal(n = length(unique(col_df$time_hr)), "Blues")
names(time_colors) <- levels(factor(col_df$time_hr))

# put colors in list
ann_colors <- list(cell_line = cell_colors, time_hr = time_colors)

# extract only significant genes and REs ---------------------------------------
sig_genes <- df %>% filter(feature_id %in% cds$gene_name & FDR < sig) %>% pull(feature_id)
sig_res <- df %>% filter(!feature_id %in% cds$gene_name & FDR < sig) %>% pull(feature_id)
lcpms.sig_genes <- subset(lcpms.gene, rownames(lcpms.gene) %in% sig_genes)
lcpms.sig_res <- subset(lcpms.re, rownames(lcpms.re) %in% sig_res)

# plot heatmaps ----------------------------------------------------------------
quickmap(lcpms.sig_genes,
         annotation_col = col_df,
         annotation_colors = ann_colors,
         main = paste("Significant (FDR <", sig, ") Protein Coding Genes"),
         filename = here("results", "PRJNA471642", "figures", "gene-heatmap.png"))

quickmap(lcpms.sig_res,
         annotation_col = col_df,
         annotation_colors = ann_colors,
         main = paste("Significant (FDR <", sig, ") Repetitive Elements"),
         filename = here("results", "PRJNA471642", "figures", "re-heatmap.png"))
```

## Create family level dotplots for RE

```{r eval=FALSE, include=FALSE}
d1 <- plot_fam_counts(dre, "SD6.125nm.6hr")
d2 <- plot_fam_counts(dre, "SD6.125nm.24hr")

(d1 | d2) +
  plot_annotation("Family-level RE Expression") +
  ggsave(here("results", "PRJNA471642", "figures", "dotplots.png"),
         width = 11,
         height = 7)
```

