---
title: "PRJNA391270 Analysis"
output:
  html_document:
    df_print: paged
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
source("helpers/helpers.R")
```

## Load Libraries

```{r}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(MultiAssayExperiment))
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(qsmooth))
suppressPackageStartupMessages(library(quantro))
suppressPackageStartupMessages(library(doParallel))
registerDoParallel(cores = 12)
```

## Create results directory for project if one doesn't exist

```{r}
if (!dir.exists(here("results", "PRJNA391270"))) {
  dir.create(here("results", "PRJNA391270", "rds-files"), recursive = TRUE)
  dir.create(here("results", "PRJNA391270", "data-files"))
}
```

## Read in MultiAssayExperiment Object

```{r}
mae <- readRDS(here("data", "PRJNA391270", "MultiAssayExperiment.rds"))
```

## Read in gene annotations

```{r}
cds <- readRDS(here("doc", "gencode.v26.annotation.cds.rds"))
```

## Extract RE and Gene Counts as matrices

```{r}
genes <- assay(mae, "gene counts")
repeats <- assay(mae, "RE counts")
```

## Restrict REs to five classes

LINE, SINE, long terminal repeats (LTR), SVA, and DNA transposons

```{r}
repeats <- subset(repeats, grepl(rownames(repeats), pattern = "^LINE\\.|^SINE\\.|^LTR\\.|.*\\.SVA\\..*|^DNA\\."))
```

## Restrict genes to protein coding

```{r}
genes <- subset(genes, rownames(genes) %in% cds$gene_name)
```

## Combine both matrices into a single count matrix

```{r}
# check for consistent colnames
all(colnames(genes) == colnames(repeats))

# bind both matrices into single count matrix
counts <- rbind(genes, repeats)
```

## Sum technical replicates across runs and collapse metadata (if present)

Also has the effect of renaming the rows by the BioSample accession.

```{r}
counts <- sumTechReps(counts, ID = colData(mae)$BioSample)
colData(mae)$bases <- as.integer(colData(mae)$bases)
metadata <- as.data.table(colData(mae))
metadata <- unique(metadata[, .(BioSample, group)])
setDF(metadata, rownames = metadata$BioSample)

# Add additional metadata
splits <- tstrsplit(metadata$group, "\\.")
metadata$cell_line <- splits[[1]]
metadata$treatment <- splits[[2]]
```

## Check for global differences between groups with `quantro`

```{r}
qtest.cell_line <- quantro(counts, groupFactor = metadata$cell_line, B = 1e3)
perm_pval.cell_line <- quantroPvalPerm(qtest.cell_line)
qtest.treatment <- quantro(counts, groupFactor = metadata$treatment, B = 1e3)
perm_pval.treatment <- quantroPvalPerm(qtest.treatment)
perm_pval <- min(c(perm_pval.cell_line, perm_pval.treatment))
```

## DE pipeline

```{r}
# Define experimental design
design <- model.matrix(~0 + group, data = metadata)
colnames(design) <- gsub(pattern = "^group", replacement = "", x = colnames(design))

# Define contrasts to test
contrast.matrix <- makeContrasts(
  A498.4GY_vs_A498.Control = A498.4GY_3hr - A498.control,
  A549.4GY_vs_A549.Control = A549.4GY_3hr - A549.control,
  A549.Nutlin_vs_A549.Control = A549.NUT - A549.control,
  A549.TGFB_vs_A549.Control = A549.TGFB - A549.control,
  A549.TGFB_Nutlin_vs_A549.Control = A549.TGFB_NUT - A549.control,
  H460.4GY_vs_H460.Control = H460.4GY_3hr - H460.control,
  HCT116.4GY_vs_HCT116.Control = HCT116.4GY_3hr - HCT116.control,
  LOXIMVI.4GY_vs_LOXIMVI.Control = LOXIMVI.4GY_3hr - LOXIMVI.control,
  MALME3.4GY_vs_MALME3.Control = MALME3.4GY_3hr - MALME3.control,
  MCF7.4GY_vs_MCF7.Control = MCF7.4GY_3hr - MCF7.control,
  SKMEL5.4GY_vs_SKMEL5.Control = SKMEL5.4GY_3hr - SKMEL5.control,
  U2OS.4GY_vs_U2OS.Control = U2OS.4GY_3hr - U2OS.control,
  UACC257.4GY_vs_UACC257.Control = UACC257.4GY_3hr - UACC257.control,
  UO31.4GY_vs_UO31.Control = UO31.4GY_3hr - UO31.control,
  
  x4GY_vs_Control = ((A498.4GY_3hr + 
                        A549.4GY_3hr + 
                        H460.4GY_3hr + 
                        HCT116.4GY_3hr + 
                        LOXIMVI.4GY_3hr +
                        MALME3.4GY_3hr +
                        MCF7.4GY_3hr + 
                        SKMEL5.4GY_3hr +
                        U2OS.4GY_3hr +
                        UACC257.4GY_3hr +
                        UO31.4GY_3hr) / 11) - 
    ((A498.control +
        A549.control +
        H460.control +
        HCT116.control +
        LOXIMVI.control +
        MALME3.control +
        MCF7.control +
        SKMEL5.control +
        U2OS.control +
        UACC257.control +
        UO31.control) / 11),
  levels = design
)

# Create DGEList object from counts and filter for expression
y <- DGEList(counts = counts, samples = metadata)
keep <- filterByExpr(y, design = design)
y <- y[keep,, keep.lib.sizes = FALSE]

# If global differences exist use QSmooth normalization, if not use TMM
if (perm_pval < 0.01) {
  norm_method <- "qs"
  qs <- qsmooth(y$counts, group_factor = y$samples$group)
  fit <- lmFit(log2(qsmoothData(qs) + 2), design)
  fit <- contrasts.fit(fit, contrast.matrix)
} else {
  norm_method <- "tmm"
  y <- calcNormFactors(y, method = "TMM")
  y <- estimateDisp(y, design, robust = TRUE)
  fit <- glmQLFit(y, design, robust = TRUE)
  plotQLDisp(fit)
}
```

## Test for differential expression above a fold-change

Use either `treat` or `glmTreat` depending on the fit type to test for 
differential expression above a fold-change.

```{r}
# Define fold-change that features must exceed to be considered for significance
FC <- 1.5

contrast_names <- colnames(contrast.matrix)
if (norm_method == "qs") {
  fit2 <- treat(fit, lfc = log2(FC), trend = TRUE, robust = TRUE)
  plotSA(fit2)
  res_dts <- get_treat(contrast_names, fit2)
} else {
  res_dts <- get_glmTreat(contrast_names, fit, contrast.matrix, FC)
  }

# extract all results and bind to single df
df <- rbindlist(res_dts, idcol = "contrast")
```

## Save a `SummarizedExperiment` of processed data

Save the:

- Filtered raw counts (gene + RE)
- Metadata used for DE testing
- Design matrix used for DE testing
- Model fit used in DE testing
- Differential expression results

```{r}
se <- SummarizedExperiment(assays = list("counts" = y$counts),
                           colData = metadata,
                           metadata = list("design" = design, 
                                           "contrasts" = contrast.matrix,
                                           "fit" = fit,
                                           "de" = df))

saveRDS(se, here("results", "PRJNA391270", "rds-files", "se.rds"))
```
