---
title: "PRJNA523616 Analysis"
output:
  html_document:
    df_print: paged
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
source("helpers/helpers.R")
```

## Read in MultiAssayExperiment Object

```{r}
library(here)
library(data.table)
library(MultiAssayExperiment)


mae <- readRDS(here("data", "PRJNA523616", "MultiAssayExperiment.rds"))
```

## Read in gene annotations

```{r}
cds <- readRDS(here("doc", "gencode.v26.annotation.cds.rds"))
```

## Extract RE and Gene Counts as matrices

```{r}
genes <- assay(mae, "gene counts")
repeats <- assay(mae, "RE counts")
```

## Restrict REs to five classes

LINE, SINE, long terminal repeats (LTR), SVA, and DNA transposons

```{r}
repeats <- subset(repeats, grepl(rownames(repeats), pattern = "^LINE\\.|^SINE\\.|^LTR\\.|.*\\.SVA\\..*|^DNA\\."))
```

## Restrict genes to protein coding

```{r}
genes <- subset(genes, rownames(genes) %in% cds$gene_name)
```

## Combine both matrices into a single count matrix

```{r}
# check for consistent colnames
all(colnames(genes) == colnames(repeats))

# bind both matrices into single count matrix
counts <- rbind(genes, repeats)
```

## Sum technical replicates across runs and collapse metadata (if present)

Also has the effect of renaming the rows by the bio_sample accession.

```{r}
counts <- edgeR::sumTechReps(counts, ID = colData(mae)$bio_sample)

metadata <- as.data.table(colData(mae))
metadata <- unique(metadata[, .(bio_sample, group)])
setDF(metadata, rownames = metadata$bio_sample)
all(colnames(counts) == rownames(metadata))
```

## DE pipeline

```{r}
design <- model.matrix(~0 + group, data = metadata)
colnames(design) <- gsub(pattern = "^group", replacement = "", x = colnames(design))
y <- DGEList(counts = counts, samples = metadata)
keep <- filterByExpr(y, design = design)
y <- y[keep,, keep.lib.sizes = FALSE]
y <- calcNormFactors(y, design = design, method = "TMM")
y <- estimateDisp(y, design = design, robust = TRUE)
fit <- glmQLFit(y, design = design, robust = TRUE)
```

## DE testing

```{r}
contrast.matrix <- makeContrasts(
  F36P.Etoposide_vs_Untreated = F36P.Etoposide - F36P.Untreated,
  HL60.Etoposide_vs_untreated = HL60.Etoposide - HL60.Untreated,
  KASUMI1.Etoposide_vs_Untreated = KASUMI1.Etoposide - KASUMI1.Untreated,
  MOLM13.Etoposide_vs_untreated = MOLM13.Etoposide - MOLM13.Untreated,
  MONOMAC6.Etoposide_vs_Untreated = MONOMAC6.Etoposide - MONOMAC6.Untreated,
  MV411.Etoposide_vs_Untreated = MV411.Etoposide - MV411.Untreated,
  NB4.Etoposide_vs_Untreated = NB4.Etoposide - NB4.Untreated,
  NOMO1.Etoposide_vs_Untreated = NOMO1.Etoposide - NOMO1.Untreated,
  OCIAML2.Etoposide_vs_Untreated = OCIAML2.Etoposide - OCIAML2.Untreated,
  OCIAML3.Etoposide_vs_Untreated = OCIAML3.Etoposide - OCIAML3.Untreated,
  THP1.Etoposide_vs_Untreated = THP1.Etoposide - THP1.Untreated,
  
  levels = design
)

# create inputs to map over (titles are same as above)
cons <- colnames(contrast.matrix)
names(cons) <- cons

# set global significance levels (used in all downstream plots/analyses)
fc <- 1.5

# map over and extract results
de_results <- lapply(
  cons, 
  get_de_results2, 
  glm_fit = fit, 
  contrast_matrix = contrast.matrix,
  fc = fc
)

# extract all results and bind to single df
df <- rbindlist(lapply(de_results, function(x) x[["table"]]), idcol = "contrast")
```

## Create results directory for project

```{r}
if (!dir.exists(here("results", "PRJNA523616"))) {
  dir.create(here("results", "PRJNA523616", "rds-files"), recursive = TRUE)
  dir.create(here("results", "PRJNA523616", "data-files"))
}
```

## Write results out

```{r}
# split DE results into gene and RE results and write out
fwrite(
  x = df[feature_id %chin% cds$gene_name],
  file = here("results", "PRJNA523616", "data-files", "dge.tsv"),
  sep = "\t"
  )

fwrite(
  x = df[!feature_id %chin% cds$gene_name],
  file = here("results", "PRJNA523616", "data-files", "dre.tsv"),
  sep = "\t"
)
```

## Calculate normalized counts

```{r}
lcpms <- cpm(y, log = TRUE, prior.count = 2)

# split lcpms by gene and by RE
lcpms.gene <- subset(lcpms, rownames(lcpms) %in% cds$gene_name)
lcpms.re <- subset(lcpms, !rownames(lcpms) %in% cds$gene_name)

# Write normalized counts out
saveRDS(lcpms.gene, here("results", "PRJNA523616", "rds-files", "gene-lcpms.rds"))
saveRDS(lcpms.re, here("results", "PRJNA523616", "rds-files", "re-lcpms.rds"))
```

## Save design and contrast matrix

```{r}
saveRDS(design, here("results", "PRJNA523616", "rds-files", "design.rds"))
saveRDS(contrast.matrix, here("results", "PRJNA523616", "rds-files", "contrast-matrix.rds"))
```
