---
title: "PRJNA607909 Analysis"
output:
  html_document:
    df_print: paged
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 4
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

source("helpers/helpers.R")
```

## Read in SummarizedExperiment Object

```{r}
library(tidyverse)
library(here)
library(SummarizedExperiment)


se <- readRDS(here("data", "PRJNA607909", "summarized_experiment.rds"))
```

## Read in gene annotations

```{r}
cds <- read_rds(here("doc", "gencode.v26.annotation.cds.rds"))
```

## Filter out RNA REs from counts assay

```{r}
se <- subset(se, !str_detect(rownames(se), "tRNA|snRNA|scRNA|rRNA|srpRNA|Simple_repeat|Low_complexity"))
```

## DE pipeline

The design of this experiment is somewhat different from the others. There is 
only on replicate for each cell_type so we can't properly test for differences
within cell_types. However we can test DAC vs DMSO controlling for cell_type.

```{r}
library(edgeR)
library(coriell)


Treatment <- factor(colData(se)$treatment, levels = c("pbs", "dac"))
Cell_type <- factor(colData(se)$cell_type)
design <- model.matrix(~Cell_type + Treatment)
y <- DGEList(counts = assay(se, "counts"),
             samples = colData(se))
keep <- filterByExpr(y, design = design)
y <- y[keep,, keep.lib.sizes = FALSE]
y <- calcNormFactors(y, design = design, method = "TMM")
y <- estimateDisp(y, design = design, robust = TRUE)
fit <- glmQLFit(y, design = design, robust = TRUE)
```

## DE testing

```{r}
library(patchwork)

# define global significant cutoffs
fc = 1.5
sig = 0.1

# test for overall differences 
dac_vs_dmso <- glmTreat(fit, lfc = log2(fc))

# volcano plots
v1 <- edger_to_df(dac_vs_dmso) %>%
  filter(feature_id %in% cds$gene_name) %>% 
  plot_volcano(fdr = sig, lfc = log2(fc)) +
  ggtitle("Protein Coding Gene Expression")

v2 <- edger_to_df(dac_vs_dmso) %>%
  filter(!feature_id %in% cds$gene_name) %>% 
  plot_volcano(fdr = sig, lfc = log2(fc)) +
  ggtitle("RE Expression")

m1 <- edger_to_df(dac_vs_dmso) %>% 
  filter(feature_id %in% cds$gene_name) %>% 
  plot_md(fdr = sig, lfc = log2(fc)) +
  ggtitle("Protein Coding Gene Expression")

m2 <- edger_to_df(dac_vs_dmso) %>% 
  filter(!feature_id %in% cds$gene_name) %>% 
  plot_md(fdr = sig, lfc = log2(fc)) +
  ggtitle("RE Expression")

# create patchwork
(v1 | v2) / (m1 | m2) +
plot_annotation("DAC vs DMSO (controlling for Cell Type Differences)") +
plot_layout(guides = "collect") +
ggsave(here("results", "PRJNA607909", "figures", "patchwork.png"),
       width = 11,
       height = 7)
```

## Write results out

```{r}
# split DE results into gene and RE results and write out
dge <- edger_to_df(dac_vs_dmso) %>% 
  filter(feature_id %in% cds$gene_name) %>% 
  write_tsv(here("results", "PRJNA607909", "data-files", "dge.tsv"))

dre <- edger_to_df(dac_vs_dmso) %>% 
  filter(!feature_id %in% cds$gene_name) %>% 
  write_tsv(here("results", "PRJNA607909", "data-files", "dre.tsv"))
```

## Calculate normalized counts

```{r}
lcpms <- cpm(y, log = TRUE, prior.count = 2)

# split lcpms by gene and by RE
lcpms.gene <- subset(lcpms, rownames(lcpms) %in% cds$gene_name)
lcpms.re <- subset(lcpms, !rownames(lcpms) %in% cds$gene_name)

# Write normalized counts out
write_rds(lcpms.gene, here("results", "PRJNA607909", "rds-files", "gene-lcpms.rds"))
write_rds(lcpms.re, here("results", "PRJNA607909", "rds-files", "re-lcpms.rds"))
```

## PCA

```{r}
library(FactoMineR)
library(factoextra)


# perform PCA with factominer --------------------------------------------------
pca.gene <- PCA(t(lcpms.gene),
                scale.unit = TRUE,
                graph = FALSE,
                ncp = 5)

pca.re <- PCA(t(lcpms.re),
              scale.unit = TRUE,
              graph = FALSE,
              ncp = 5)

# scree plot -------------------------------------------------------------------
fviz_eig(pca.gene, addlabels = TRUE) +
  ggtitle("Protein Coding Genes") +
  ggsave(here("results", "PRJNA607909", "figures", "gene-scree.png"),
         width = 8,
         height = 5)

fviz_eig(pca.re, addlabels = TRUE) +
  ggtitle("Repetitive Elements") +
  ggsave(here("results", "PRJNA607909", "figures", "re-scree.png"),
         width = 8,
         height = 5)

# extract the individual level coordinates -------------------------------------
ind.gene <- get_pca_ind(pca.gene)
ind.re <- get_pca_ind(pca.re)

# plot -------------------------------------------------------------------------
# genes
ind.gene$coord %>% 
  as_tibble(rownames = "sample_name") %>% 
  left_join(as_tibble(colData(se), rownames = "sample_name"), by = "sample_name") %>% 
  ggplot(aes(x = Dim.1, y = Dim.2, color = cell_type, shape = treatment)) +
  geom_point(size = 5) +
  geom_hline(yintercept = 0, linetype = 3) +
  geom_vline(xintercept = 0, linetype = 3) +
  labs(title = "Protein Coding Genes",
       x = "PC 1",
       y = "PC 2") +
  scale_color_brewer(palette = "Dark2") +
  theme_light() +
  ggsave(here("results", "PRJNA607909", "figures", "gene-pca.png"),
         width = 8,
         height = 5)
# REs
ind.re$coord %>% 
  as_tibble(rownames = "sample_name") %>% 
  left_join(as_tibble(colData(se), rownames = "sample_name"), by = "sample_name") %>% 
  ggplot(aes(x = Dim.1, y = Dim.2, color = cell_type, shape = treatment)) +
  geom_point(size = 5) +
  geom_hline(yintercept = 0, linetype = 3) +
  geom_vline(xintercept = 0, linetype = 3) +
  labs(title = "Repetitive Elements",
       x = "PC 1",
       y = "PC 2") +
  scale_color_brewer(palette = "Dark2") +
  theme_light() +
  ggsave(here("results", "PRJNA607909", "figures", "re-pca.png"),
         width = 8,
         height = 5)
```

## Heatmaps

```{r}
# create column annotation
col_df <- colData(se)[, c("cell_type", "treatment")] %>%
  as.data.frame() %>% 
  mutate(cell_type = factor(cell_type),
         treatment = factor(treatment))

# create colors for column annotations
trt_colors <-  c("#377EB8", "#E41A1C")
cell_colors <- RColorBrewer::brewer.pal(n = nlevels(col_df$cell_type), "Dark2")
names(cell_colors) <- levels(col_df$cell_type)
names(trt_colors) <- levels(col_df$treatment)

# put colors in list
ann_colors <- list(cell_type = cell_colors, treatment = trt_colors)

# extract only significant genes and REs ---------------------------------------
sig_genes <- edger_to_df(dac_vs_dmso) %>% filter(feature_id %in% cds$gene_name & FDR < sig) %>% pull(feature_id)
sig_res <- edger_to_df(dac_vs_dmso) %>% filter(!feature_id %in% cds$gene_name & FDR < sig) %>% pull(feature_id)
lcpms.sig_genes <- subset(lcpms.gene, rownames(lcpms.gene) %in% sig_genes)
lcpms.sig_res <- subset(lcpms.re, rownames(lcpms.re) %in% sig_res)

# plot heatmaps ----------------------------------------------------------------
quickmap(lcpms.sig_genes, 
         annotation_col = col_df,
         annotation_colors = ann_colors,
         main = paste("Significant (FDR <", sig, ") Protein Coding Genes"),
         filename = here("results", "PRJNA607909", "figures", "gene-heatmap.png"))

quickmap(lcpms.sig_res, 
         annotation_col = col_df,
         annotation_colors = ann_colors,
         main = paste("Significant (FDR <", sig, ") Repetitive Elements"),
         filename = here("results", "PRJNA607909", "figures", "re-heatmap.png"))
```

## Create Family-Level Dotplots for significant REs

```{r}
de_df <- edger_to_df(dac_vs_dmso) %>% 
  filter(!feature_id %in% cds$gene_name) %>% 
  mutate(contrast = "dac_vs_dmso")

plot_fam_counts(de_df, "dac_vs_dmso") +
  ggsave(here("results", "PRJNA607909", "figures", "dotplot.png"),
         width = 8,
         height = 5)
```
